---
title: "Genomic Annotation of Microsatellites (STRs)"
author: "Rodrigues-Silva A."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

# Methodological Note

This pipeline implements `locateVariants()` from the `VariantAnnotation`
package (Obenchain et al., *Bioinformatics* 2014) to determine the
**genomic location** of microsatellite (STR) loci. The approach is
analogous to the one used by Han et al. (*PLOS Genetics* 2022) for
*Plasmodium*.
**Important:** this analysis does **not** infer functional consequence
or protein impact — it is limited to classifying loci by genomic region
(intron, UTR, CDS, intergenic, splice site, promoter).

This pipeline **does not call** `predictCoding()` or `summarizeVariants()`,
since those functions assume point variants in CDS and are not valid for
microsatellites. Localization via `locateVariants()` is sufficient and
methodologically supported (Han et al. 2022; Obenchain et al. 2014).
STRs classified as `coding` indicate overlap with CDS, but **do not
imply frameshift** — that depends on the repeat motif and requires
specific tools (HipSTR, GangSTR).

**Note on counting units:** throughout this report, two types of counts
appear side by side and **are not interchangeable**:

- **Annotations**: each row generated by `locateVariants()`. A single STR
  can produce multiple annotations if it overlaps more than one transcript
  (e.g., alternative isoforms of the same gene, or neighboring genes in
  the case of intergenic regions).
- **Unique STRs**: each STR counted at most once per region category,
  after deduplication by genomic identifier (`STR_UID`).

Section 4.1 reports raw annotations; Sections 5, 6, 7, and 10 report
unique STRs, unless stated otherwise.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.align = "center"
)
```

---

# 1. Setup {.tabset}

## 1.1 Packages

```{r packages}
library(VariantAnnotation)
library(GenomicFeatures)
library(txdbmaker)
library(GenomicRanges)
library(GenomeInfoDb)
library(IRanges)
library(Biostrings)
library(rtracklayer)
library(Rsamtools)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(ggplot2)
library(scales)
library(patchwork)
library(DT)
library(kableExtra)
library(readr)
library(writexl)
library(ggthemes)
library(vcfR)
```

## 1.2 Configuration

```{r config}
VCF_STR   <- "hipstr_trf_v2_filtered.vcf"
GENOME_FA <- "genome.fasta"
GTF_FILE  <- "mikado.loci.clean.step2.primaryTranscripts.gff3"
PROM_UP   <- 2000
PROM_DOWN <- 200
OUT_DIR   <- "str_results"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
OUT_TSV   <- file.path(OUT_DIR, "str_localized.tsv")
OUT_EXCEL <- file.path(OUT_DIR, "str_localized.xlsx")
OUT_SUMMARY_CSV <- file.path(OUT_DIR, "loc_summary.csv")
TXDB_FILE <- file.path(OUT_DIR, "txdb.sqlite")
```

# 2. Building the TxDb

```{r txdb}
build_txdb <- function(genome_fa, gtf_file, txdb_file) {
  # Index FASTA if needed
  if (!file.exists(paste0(genome_fa, ".fai")))
    indexFa(genome_fa)
  fa <- FaFile(genome_fa)
  on.exit(close(fa), add = TRUE)
  open(fa)
  fa_seqs <- seqnames(seqinfo(fa))
  # Return existing TxDb if already built
  if (file.exists(txdb_file)) {
    message("Existing TxDb found: ", txdb_file)
    return(loadDb(txdb_file))
  }
  # Import GFF/GTF
  fmt <- ifelse(grepl("\\.gff3?$", gtf_file, ignore.case = TRUE), "gff3", "gtf")
  gr_full <- rtracklayer::import(gtf_file, format = fmt)
  # Filter contigs missing from the FASTA
  gff_seqs    <- as.character(unique(seqnames(gr_full)))
  missing_seq <- setdiff(gff_seqs, fa_seqs)
  if (length(missing_seq) > 0) {
    message(
      length(missing_seq), " contigs in the GFF missing from the FASTA — removed: ",
      paste(head(missing_seq, 5), collapse = ", "),
      if (length(missing_seq) > 5) " ..."
    )
    gr_full <- gr_full[as.character(seqnames(gr_full)) %in% fa_seqs]
    seqlevels(gr_full, pruning.mode = "coarse") <-
      intersect(fa_seqs, seqlevels(gr_full))
  }
  seqinfo(gr_full) <- seqinfo(fa)[seqlevels(gr_full)]
  message("Features retained: ", length(gr_full))
  # Build and save TxDb
  txdb <- txdbmaker::makeTxDbFromGRanges(gr_full, taxonomyId = NA)
  saveDb(txdb, txdb_file)
  message("TxDb saved: ", txdb_file)
  txdb
}
txdb <- build_txdb(GENOME_FA, GTF_FILE, TXDB_FILE)
cat(sprintf(
  "Genes: %d | Transcripts: %d | Exons: %d | CDS: %d\n",
  length(genes(txdb)), length(transcripts(txdb)),
  length(exons(txdb)), length(cds(txdb))
))
```

# 3. Reading the VCF

```{r read_vcf}
vcf <- readVcf(VCF_STR, genome = "custom")
cat(sprintf("STRs: %d | Samples: %d\n", nrow(vcf), ncol(vcf)))
# Allele length summary
ref_len <- width(ref(vcf))
alt_len <- vapply(alt(vcf), function(a) max(width(a)), integer(1))
cat(sprintf(
"REF — median: %d, max: %d\nALT — median: %d, max: %d\n",
  median(ref_len), max(ref_len), median(alt_len), max(alt_len)
))
```

## 3.1 Seqlevels Harmonization

```{r seqlevels}
harmonize_seqlevels <- function(vcf, txdb) {
  vcf_seqs  <- seqlevels(vcf)
  txdb_seqs <- seqlevels(txdb)
  common    <- intersect(vcf_seqs, txdb_seqs)
  # Try removing the chr prefix if there is no direct match
  if (length(common) == 0) {
    message("No common contigs — trying chr prefix removal...")
    vcf <- renameSeqlevels(vcf, setNames(sub("^[Cc]hr", "", vcf_seqs), vcf_seqs))
    common <- intersect(seqlevels(vcf), txdb_seqs)
  }
  stopifnot("No common contigs between VCF and TxDb" = length(common) > 0)
  cat(sprintf("Common contigs: %d\n", length(common)))
  vcf <- vcf[as.character(seqnames(vcf)) %in% common]
  seqlevels(vcf, pruning.mode = "coarse") <- common
  cat(sprintf("STRs after filtering: %d\n", nrow(vcf)))
  vcf
}
vcf <- harmonize_seqlevels(vcf, txdb)
```

## 3.2 Unique STR Identifier (`STR_UID`)

`rownames(vcf)` (the VCF `ID` field) is not reliable as a deduplication
identifier: callers such as HipSTR often leave this field as `"."` for
all records, or may generate repeated values. Since any "unique STRs"
count in the following sections depends on an identifier that is truly
unique per STR, we build here an ID based on the genomic position
(`CHROM:START`), which is always unique.

```{r str_uid}
n_dup_id <- sum(duplicated(rownames(vcf)))
if (n_dup_id > 0) {
  cat(sprintf(
    "Warning: %d duplicated rownames(vcf) — STR_ID is not reliable for deduplication; using STR_UID (CHROM:START)\n",
    n_dup_id
  ))
} else {
  cat("rownames(vcf) are unique — no duplicates found.\n")
}

str_uid <- paste0(as.character(seqnames(vcf)), ":", start(vcf))
stopifnot("STR_UID contains duplicates — check the VCF (repeated positions?)" =
            !any(duplicated(str_uid)))
cat(sprintf("Unique STR_UIDs: %d (should equal the total number of STRs: %d)\n",
            length(unique(str_uid)), nrow(vcf)))
```

# 4. locateVariants() — Regional Annotation {.tabset}

`locateVariants()` maps each STR to all overlapping genomic regions.
An STR at an exon-intron junction will receive multiple annotations (expected).
| Class | Region |
|--------|--------|
| `CodingVariants()` | CDS |
| `IntronVariants()` | Intron |
| `FiveUTRVariants()` | 5' UTR |
| `ThreeUTRVariants()` | 3' UTR |
| `SpliceSiteVariants()` | ±2 bp from the exon-intron junction |
| `PromoterVariants()` | Promoter (customizable) |
| `IntergenicVariants()` | Intergenic |

## 4.1 Execution

```{r locate}
loc <- locateVariants(
  query   = vcf,
  subject = txdb,
  region  = AllVariants(
    promoter   = PromoterVariants(upstream = PROM_UP, downstream = PROM_DOWN),
    intergenic = IntergenicVariants(upstream = 1e6L, downstream = 1e6L)
  )
)
cat(sprintf(
  "Annotations (STR x transcript x region): %d | Annotated STRs: %d\n",
  length(loc), length(unique(loc$QUERYID))
))
cat("\n--- ANNOTATION distribution (not deduplicated; see Section 5.1 for unique STRs) ---\n")
print(sort(table(loc$LOCATION), decreasing = TRUE))
```

*The numbers above count **annotations**, not STRs. Since an STR can
overlap more than one transcript, it may generate multiple rows in
`loc` — one per STR × transcript × region combination. Sections 5 and 6,
in turn, report the number of **SSRs with at least 1 annotation**: each SSR counted at most once
WITHIN each region category (`LOCATION`), after deduplication by
`STR_UID`. Since the same SSR can have an annotation in more than one
category simultaneously (e.g., promoter + fiveUTR), the sum of these
counts across categories may exceed the overall total of annotated
SSRs in the dataset (see Section 5.2).*

## 4.2 Results Table

```{r build_table}
loc_df <- as.data.frame(loc, row.names = NULL) |>
  dplyr::transmute(
    CHROM    = seqnames,
    START    = start,
    END      = end,
    STR_ID   = rownames(vcf)[QUERYID],   # original VCF ID — for reference/display only
    STR_UID  = str_uid[QUERYID],          # unique identifier (CHROM:START) — use for dedup/counts
    REF      = as.character(ref(vcf)[QUERYID]),
    ALT      = vapply(alt(vcf)[QUERYID], function(a) {
      paste(as.character(a), collapse = ",")
    }, character(1)),
    REF_LEN  = nchar(REF),
    ALT_LEN  = nchar(sub(",.*", "", ALT)),
    DELTA_BP = ALT_LEN - REF_LEN,
    QUAL     = qual(vcf)[QUERYID],
    FILTER   = as.character(filt(vcf)[QUERYID]),
    GENE_ID  = as.character(GENEID),
    TX_ID    = as.character(TXID),
    LOCATION = as.character(LOCATION)
  ) |>
  dplyr::arrange(CHROM, START)
cat(sprintf("Table: %d rows (annotations) × %d columns | %d STRs (STR_UID)\n",
            nrow(loc_df), ncol(loc_df), dplyr::n_distinct(loc_df$STR_UID)))
```

# 5. Descriptive Statistics {.tabset}

All counts in this section use `STR_UID` for deduplication: each STR
is counted at most once per region category, regardless of how many
transcripts it overlaps.

## 5.1 STRs by Region

```{r stats_region}
region_unique <- loc_df |>
  dplyr::distinct(STR_UID, LOCATION) |>
  dplyr::count(LOCATION, name = "n_STRs") |>
  dplyr::arrange(desc(n_STRs)) |>
  dplyr::mutate(pct = n_STRs / sum(n_STRs))

kableExtra::kbl(
  region_unique |> mutate(pct = scales::percent(pct, 0.1)),
  col.names = c("Region", "STRs ≥ 1 annotation", "%"),
  caption   = "STRs by genomic region"
) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

## 5.2 Multi-region Overlap

```{r stats_overlap}
multi <- loc_df |>
  dplyr::group_by(STR_UID) |>
  dplyr::summarise(n_regioes = n_distinct(LOCATION), .groups = "drop") |>
  dplyr::count(n_regioes, name = "n_STRs") |>
  dplyr::mutate(pct = scales::percent(n_STRs / sum(n_STRs), 0.1))

cat(sprintf(
  "In 1 region: %d | In 2+ regions: %d\n",
  sum(multi$n_STRs[multi$n_regioes == 1]),
  sum(multi$n_STRs[multi$n_regioes > 1])
))
```

## 5.3 Motif Extraction from the INFO Field

```{r}
library(vcfR)

vcfr_obj  <- vcfR::read.vcfR(VCF_STR, verbose = FALSE)
vcfr_uid  <- paste0(vcfr_obj@fix[, "CHROM"], ":", vcfr_obj@fix[, "POS"])

stopifnot("STR_UID (vcfR) contains duplicates" = !any(duplicated(vcfr_uid)))

info_raw <- vcfr_obj@fix[, "INFO"]

period_df <- tibble::tibble(
  STR_UID = vcfr_uid,
  PERIOD  = as.integer(stringr::str_extract(info_raw, "(?<=PERIOD=)[0-9]+"))
) |>
  dplyr::mutate(
    MOTIF_CLASS = dplyr::case_when(
      PERIOD == 1 ~ "mono",
      PERIOD == 2 ~ "di",
      PERIOD == 3 ~ "tri",
      PERIOD == 4 ~ "tetra",
      PERIOD == 5 ~ "penta",
      PERIOD == 6 ~ "hexa",
      PERIOD == 7 ~ "hepta",
      PERIOD == 8 ~ "octa",
      PERIOD == 9 ~ "nona",
      TRUE ~ NA_character_
    ),
    MOTIF_CLASS = factor(MOTIF_CLASS,
                          levels = c("mono","di","tri","tetra","penta","hexa","hepta", "octa", "nona"))
  )
```


# 6. Visualizations {.tabset}

```{r palette, include=FALSE}
location_pal <- c(
  coding     = "#E6393A",
  spliceSite = "#003399",
  fiveUTR    = "#4DAF4A",
  threeUTR   = "#F5C400",
  intron     = "#984EA3",
  promoter   = "#A65628",
  intergenic = "#999999"
)

theme_publication <- function(base_size = 12, base_family = "sans") {
  theme_foundation(base_size = base_size, base_family = base_family) +
  theme(
    plot.background   = element_rect(fill = "white", colour = NA),
    panel.background  = element_rect(fill = "white", colour = NA),
    panel.border      = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    panel.grid.major  = element_line(colour = "#E8E8E8", linewidth = 0.35),
    panel.grid.minor  = element_blank(),
    axis.line         = element_blank(),
    axis.ticks        = element_line(colour = "black", linewidth = 0.4),
    axis.ticks.length = unit(2.5, "pt"),
    axis.text         = element_text(colour = "black", size = rel(0.88)),
    axis.title        = element_text(colour = "black", size = rel(0.95),
                                     margin = margin(t = 4, r = 4)),
    plot.title    = element_text(
      face = "bold", size = rel(1.08),
      colour = "black", hjust = 0,
      margin = margin(b = 3)
    ),
    plot.subtitle = element_text(
      size = rel(0.85), colour = "#444444",
      hjust = 0, margin = margin(b = 8)
    ),
    plot.caption  = element_text(
      size = rel(0.75), colour = "#666666",
      hjust = 1, margin = margin(t = 6)
    ),
    plot.margin   = margin(14, 16, 10, 12),
    legend.background  = element_rect(fill = "white", colour = NA),
    legend.key         = element_rect(fill = "white", colour = NA),
    legend.key.size    = unit(10, "pt"),
    legend.text        = element_text(size = rel(0.82)),
    legend.title       = element_text(face = "bold", size = rel(0.88)),
    legend.position    = "right",
    strip.background = element_rect(fill = "#F2F2F2", colour = "black", linewidth = 0.5),
    strip.text       = element_text(face = "bold", size = rel(0.88))
  )
}
```

## 6.1 STR Distribution by Chromosome

Unique STRs (by `STR_UID`) by chromosome and region. The sum of all
cells is checked against the total number of unique STRs from Section
5.1 — the two must match exactly, since both use the same
deduplication.

```{r chrom_summary}
loc_summary <- loc_df |>
  dplyr::distinct(CHROM, STR_UID, LOCATION) |>
  dplyr::count(CHROM, LOCATION) |>
  tidyr::pivot_wider(names_from = LOCATION, values_from = n, values_fill = 0)

# --- Sanity check: total must match the unique STRs total (5.1) ---
total_por_chrom  <- sum(dplyr::select(loc_summary, -CHROM))
total_por_regiao <- sum(region_unique$n_STRs)

cat(sprintf(
  "loc_summary total (by chromosome): %d | region_unique total (5.1): %d | %s\n",
  total_por_chrom, total_por_regiao,
  if (total_por_chrom == total_por_regiao) "OK — totals match" else "MISMATCH — investigate"
))

write.csv(loc_summary, OUT_SUMMARY_CSV, row.names = FALSE)
cat(sprintf("Saved: %s\n", OUT_SUMMARY_CSV))
```

## 6.2 Distribution by Region

```{r plot_region, fig.width=8, fig.height=5}
region_unique |>
  dplyr::mutate(LOCATION = fct_reorder(LOCATION, n_STRs)) |>
  ggplot(aes(LOCATION, n_STRs, fill = LOCATION)) +
  geom_col(
    width = 0.65,
    show.legend = FALSE,
    colour = "white",
    linewidth = 0.25
  ) +
  geom_text(
    aes(label = sprintf("%d  (%.1f%%)", n_STRs, pct * 100)),
    hjust    = -0.08,
    size     = 3.1,
    fontface = "plain",
    colour   = "black"
  ) +
  scale_fill_manual(values = location_pal) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.22)),
    labels = scales::comma
  ) +
  coord_flip() +
  labs(
    title    = "Microsatellites by Genomic Region",
    subtitle = sprintf(
      "locateVariants()  \u2022  %d SSRs with ≥ 1 annotation  \u2022  categories not mutually exclusive",
      dplyr::n_distinct(loc_df$STR_UID)
    ),
    x        = NULL,
    y        = "Number of STRs",
    caption  = "Regions ordered by ascending count"
  ) +
  theme_publication(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "#E8E8E8", linewidth = 0.35)
  )
```

## 6.2.1 Distribution by Region and Repeat Motif

```{r}
motif_all_regions <- loc_df |>
  dplyr::distinct(STR_UID, LOCATION) |>
  dplyr::left_join(period_df, by = "STR_UID") |>
  dplyr::filter(!is.na(MOTIF_CLASS)) |>
  dplyr::count(LOCATION, MOTIF_CLASS, .drop = FALSE, name = "n_STRs") |>
  dplyr::group_by(LOCATION) |>
  dplyr::mutate(pct = n_STRs / sum(n_STRs)) |>
  dplyr::ungroup()

motif_pal <- c(
  mono   = "#66C2A5",
  di     = "#FC8D62",
  tri    = "#8DA0CB",
  tetra  = "#E78AC3",
  penta  = "#A6D854",
  hexa   = "#FFD92F",
  hepta  = "#E5C494",
  octa   = "#B3B3B3",
  nona   = "darkred"
)

motif_all_regions |>
  ggplot(aes(MOTIF_CLASS, n_STRs, fill = MOTIF_CLASS)) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.25) +
  scale_fill_manual(values = motif_pal, name = "Motif class", drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08)), labels = scales::comma) +
  facet_wrap(~ LOCATION, scales = "free_y", ncol = 3) +
  labs(
    title    = "SSRs by Repeat Motif Class, across Genomic Regions",
    subtitle = "locateVariants() regions \u00d7 HipSTR PERIOD (vcfR-extracted) \u2022 SSRs with ≥ 1 annotation",
    x = NULL,
    y = "Number of STRs"
  ) +
  theme_publication(base_size = 11) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x         = element_blank(),
    axis.ticks.x         = element_blank(),
    strip.text           = element_text(face = "bold"),
    legend.position       = "right"
  )
```

## 6.3 Chromosomal Distribution

```{r plot_chrom, fig.width=11, fig.height=4.5}
loc_df |>
  dplyr::distinct(CHROM, STR_UID, LOCATION) |>
  dplyr::count(CHROM, LOCATION) |>
  dplyr::mutate(
    CHROM    = fct_reorder(CHROM, n, sum, .desc = TRUE),
    LOCATION = factor(LOCATION, levels = names(location_pal))
  ) |>
  ggplot(aes(CHROM, n, fill = LOCATION)) +
  geom_col(
    width     = 0.72,
    colour    = "white",
    linewidth = 0.25
  ) +
  scale_fill_manual(
    values = location_pal,
    name   = "Genomic region",
    guide  = guide_legend(
      keywidth  = unit(9, "pt"),
      keyheight = unit(9, "pt"),
      ncol      = 1
    )
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05)),
    labels = scales::comma
  ) +
  labs(
    title   = "SSRs with ≥ 1 annotation by Chromosome and Genomic Region",
    subtitle = sprintf(
      "locateVariants()  \u2022  %d SSRs with ≥ 1 annotation  \u2022  categories not mutually exclusive",
      dplyr::n_distinct(loc_df$STR_UID)
    ),
    x       = NULL,
    y       = "Number of STRs",
    caption = "Chromosomes ordered by descending total count."
  ) +
  theme_publication(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "right"
  )
```

# 7. Subsets of Interest

The subsets below report **unique STRs** (by `STR_UID`) in each
category — not raw rows/annotations, since the same STR could appear
more than once in the same category if it overlaps multiple
transcripts of that region.

```{r subsets}
subsets <- list(
  "Promoter + 5'UTR" = loc_df |> filter(LOCATION %in% c("promoter", "fiveUTR")),
  "Splice Site"      = loc_df |> filter(LOCATION == "spliceSite"),
  "3' UTR"           = loc_df |> filter(LOCATION == "threeUTR")
)

n_unique_str <- vapply(subsets, function(df) dplyr::n_distinct(df$STR_UID), integer(1))

cat(paste(
  sprintf("  %s: %d unique STRs (%d annotations)",
          names(subsets), n_unique_str, vapply(subsets, nrow, integer(1))),
  collapse = "\n"
))
```

```{r splice_table}
splice_unique <- subsets[["Splice Site"]] |>
  dplyr::distinct(STR_UID, .keep_all = TRUE)

if (nrow(splice_unique) > 0) {
  kableExtra::kbl(
    head(splice_unique, 20),
    caption = "Unique STRs at splice sites (top 20)"
  ) |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = TRUE
    ) |>
    kableExtra::row_spec(0, bold = TRUE, background = "#ff7f00", color = "white")
} else {
  cat("No STR detected at splice sites.\n")
}
```

# 8. Export

```{r export}
# Full TSV (all annotations — one row per STR x transcript x region)
readr::write_tsv(loc_df, OUT_TSV)

# Excel with sheets by region (also at annotation level, not deduplicated)
region_sheets <- split(loc_df, loc_df$LOCATION)
region_sheets <- c(list(All = loc_df), region_sheets)
writexl::write_xlsx(region_sheets, OUT_EXCEL)

cat(sprintf(
  "Exported:\n  • %s\n  • %s\n  • %s\n  • %s\n",
  OUT_TSV, OUT_EXCEL, TXDB_FILE, OUT_SUMMARY_CSV
))
```

# 9. Interactive Table

*Note: the table below shows raw annotations (one row per STR ×
transcript × region). Use the filter to inspect a specific `STR_UID`
if you want to see all annotations for the same STR.*

```{r datatable}
DT::datatable(
  loc_df,
  filter     = "top",
  extensions = "Buttons",
  options    = list(
    pageLength = 20,
    scrollX    = TRUE,
    dom        = "Bfrtip",
    buttons    = c("copy", "csv", "excel")
  ),
  caption = sprintf(
    "STRs by genomic region (raw annotations) | Promoter: -%d/+%d bp from TSS",
    PROM_UP, PROM_DOWN
  )
)
```

# 10. Summary

```{r summary}
total_str_unique <- dplyr::n_distinct(loc_df$STR_UID)
total_ann         <- nrow(loc_df)

loc_table_ann    <- sort(table(loc_df$LOCATION), decreasing = TRUE)
loc_table_unique <- loc_df |>
  dplyr::distinct(STR_UID, LOCATION) |>
  dplyr::count(LOCATION, sort = TRUE)

cat(sprintf("
  SUMMARY — GENOMIC LOCATION OF STRs

  SSRs with ≥1 annotation : %d
  Total annotations     : %d
  (an STR can map to multiple regions
   and/or generate multiple annotations if it overlaps
   more than one transcript)

  Intergenic (annotations) : %s
  Intronic   (annotations) : %s
  Intergenic (unique STRs): %s
  Intronic   (unique STRs): %s

",
  total_str_unique, total_ann,
  scales::percent(sum(loc_df$LOCATION == "intergenic") / total_ann, 0.1),
  scales::percent(sum(loc_df$LOCATION == "intron") / total_ann, 0.1),
  scales::percent(loc_table_unique$n[loc_table_unique$LOCATION == "intergenic"] / total_str_unique, 0.1),
  scales::percent(loc_table_unique$n[loc_table_unique$LOCATION == "intron"] / total_str_unique, 0.1)
))

cat("\n ANNOTATION Distribution \n")
print(loc_table_ann)

cat("\n STR Distribution \n")
print(loc_table_unique)
```
