---
title: "Characterization of SSRs Discovered via TRF"
author: "Rodrigues-Silva A."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.align = "center"
)
```

---

# Artur Silva - PPGGMP (27/04/2026)

> Required packages

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(ggrepel)
library(tidyverse) 
```

# 1. Read the data

```{r}
str_data <- read_tsv("str_final.tsv", 
                     col_names = c("Chromosome", "Start", "End", "Period_size", 
                                   "Copy_number", "Pct_matches", "Pct_indels"),
                     show_col_types = FALSE)

chr_sizes <- read_tsv("chromosome_sizes.txt", 
                      col_names = c("Chromosome", "Chr_size_bp"),
                      show_col_types = FALSE)
```

# 2. Genome Size (characterized chromosomes - does not include contigs)

```{r}
TOTAL_CHR_SIZE_BP <- sum(chr_sizes$Chr_size_bp)
```


# 3. Calculate the length of each STR

```{r}
str_data <- str_data %>%
  mutate(STR_length = End - Start + 1)
```

# 4. Classify STRs by repeat type (mono, di, tri, etc.)

```{r}
str_data <- str_data %>%
  mutate(STR_type = case_when(
    Period_size == 1 ~ "Mono",
    Period_size == 2 ~ "Di",
    Period_size == 3 ~ "Tri",
    Period_size == 4 ~ "Tetra",
    Period_size == 5 ~ "Penta",
    Period_size == 6 ~ "Hexa",
    Period_size == 7 ~ "Hepta",
    Period_size == 8 ~ "Octa",
    Period_size == 9 ~ "Nona",
    TRUE ~ "Other"
  ))
```

# 5. Calculate statistics per chromosome

```{r}
stats_per_chr <- str_data %>%
  group_by(Chromosome) %>%
  summarise(
    No_of_SSRs = n(),
    Mono = sum(STR_type == "Mono"),
    Di = sum(STR_type == "Di"),
    Tri = sum(STR_type == "Tri"),
    Tetra = sum(STR_type == "Tetra"),
    Penta = sum(STR_type == "Penta"),
    Hexa = sum(STR_type == "Hexa"),
    Hepta = sum(STR_type == "Hepta"),
    Octa = sum(STR_type == "Octa"),
    Nona = sum(STR_type == "Nona"),
    Total_SSR_length_bp = sum(STR_length),
    Mean_STR_length_bp = round(mean(STR_length), 2),
    Mean_copy_number = round(mean(Copy_number), 2),
    Median_copy_number = round(median(Copy_number), 2),
    Mean_pct_matches = round(mean(Pct_matches), 2),
    .groups = 'drop'
  )
```

# 6. Join with chromosome sizes

```{r}
final_table <- chr_sizes %>%
  left_join(stats_per_chr, by = "Chromosome") %>%
  mutate(
    SSRs_per_Mb = round(No_of_SSRs / (Chr_size_bp / 1e6), 2),
    Pct_chr_in_SSRs = round((Total_SSR_length_bp / Chr_size_bp) * 100, 4)
  ) %>%
  dplyr::select(
    Chromosome,
    Chr_size_bp,
    No_of_SSRs,
    SSRs_per_Mb,
    Mono, Di, Tri, Tetra, Penta, Hexa, Hepta, Octa, Nona,
    Total_SSR_length_bp,
    Pct_chr_in_SSRs,  # Renomeado para deixar claro que é % do cromossomo
    Mean_STR_length_bp,
    Mean_copy_number,
    Median_copy_number,
    Mean_pct_matches
  ) %>%
  arrange(Chromosome)
```

# 7. Add total/mean row

```{r}
totals <- data.frame(
  Chromosome = "Total/Mean",
  Chr_size_bp = sum(final_table$Chr_size_bp),
  No_of_SSRs = sum(final_table$No_of_SSRs),
  SSRs_per_Mb = round(sum(final_table$No_of_SSRs) / (TOTAL_CHR_SIZE_BP / 1e6), 2),
  Mono = sum(final_table$Mono),
  Di = sum(final_table$Di),
  Tri = sum(final_table$Tri),
  Tetra = sum(final_table$Tetra),
  Penta = sum(final_table$Penta),
  Hexa = sum(final_table$Hexa),
  Hepta = sum(final_table$Hepta),
  Octa = sum(final_table$Octa),
  Nona = sum(final_table$Nona),
  Total_SSR_length_bp = sum(final_table$Total_SSR_length_bp),
  Pct_chr_in_SSRs = round((sum(final_table$Total_SSR_length_bp) / TOTAL_CHR_SIZE_BP) * 100, 4),
  Mean_STR_length_bp = round(mean(final_table$Mean_STR_length_bp), 2),
  Mean_copy_number = round(mean(final_table$Mean_copy_number), 2),
  Median_copy_number = round(median(final_table$Median_copy_number), 2),
  Mean_pct_matches = round(mean(final_table$Mean_pct_matches), 2)
)

final_table_with_totals <- bind_rows(final_table, totals)
```

# 8. Save results

```{r}
write_tsv(final_table_with_totals, "str_statistics_per_chromosome.tsv")
write_csv(final_table_with_totals, "str_statistics_per_chromosome.csv")
```

# 9. Display table

```{r}
cat("\n=== STR Statistics Per Chromosome ===\n\n")
print(final_table_with_totals, n = Inf)
```

# 10. General summary

```{r}
cat("\n=== Summary ===\n")
cat(sprintf("Total chromosomes analyzed: %d\n", nrow(final_table)))
cat(sprintf("Total chromosome size: %s bp (%.2f Mb)\n", 
            format(TOTAL_CHR_SIZE_BP, big.mark = ","),
            TOTAL_CHR_SIZE_BP / 1e6))
cat("\n")
cat(sprintf("Total SSRs found: %s\n", format(sum(final_table$No_of_SSRs), big.mark = ",")))
cat(sprintf("Average SSRs per Mb: %.2f\n", 
            sum(final_table$No_of_SSRs) / (TOTAL_CHR_SIZE_BP / 1e6)))
cat(sprintf("Total SSR length: %s bp\n", format(sum(final_table$Total_SSR_length_bp), big.mark = ",")))
cat(sprintf("Percentage of genome in SSRs: %.4f%%\n", 
            (sum(final_table$Total_SSR_length_bp) / TOTAL_CHR_SIZE_BP) * 100))

cat("\nResults saved to:\n")
cat("  - str_statistics_per_chromosome.tsv\n")
cat("  - str_statistics_per_chromosome.csv\n")
```

***

# 11. Pie chart: motif distribution

```{r}
# Processing data 
motif_counts <- data.frame(
  Motif = c("Mono", "Di", "Tri", "Tetra", "Penta", "Hexa", "Hepta", "Octa", "Nona"),
  Count = c(
    sum(final_table$Mono), sum(final_table$Di),   sum(final_table$Tri),
    sum(final_table$Tetra), sum(final_table$Penta), sum(final_table$Hexa),
    sum(final_table$Hepta), sum(final_table$Octa),  sum(final_table$Nona)
  )
) %>%
  filter(Count > 0) %>%
  mutate(
    Percentage = round((Count / sum(Count)) * 100, 2),
    Label      = paste0(Motif, "\n", format(Count, big.mark = ","),
                        "\n(", Percentage, "%)")
  ) %>%
  mutate(Motif = factor(Motif, levels = c("Mono", "Di", "Tri", "Tetra", "Penta",
                                           "Hexa", "Hepta", "Octa", "Nona")))

# Label positions 
motif_pos <- motif_counts %>%
  mutate(
    csum = rev(cumsum(rev(Count))),
    pos  = Count / 2 + lead(csum, 1),
    pos  = if_else(is.na(pos), Count / 2, pos)
  )

# Okabe-Ito palette 
motif_colors <- c(
  "Mono"  = "#56B4E9", "Di"    = "#0072B2", "Tri"   = "#009E73",
  "Tetra" = "#F0E442", "Penta" = "#E69F00", "Hexa"  = "#D55E00",
  "Hepta" = "#CC79A7", "Octa"  = "#999999", "Nona"  = "#2C2C2A"
)

# Legend with n and % 
legend_labels <- setNames(
  paste0(
    formatC(as.character(motif_counts$Motif), width = 5, flag = "-"),
    "  n=", formatC(motif_counts$Count, format = "d", big.mark = ",", width = 7),
    "  (", motif_counts$Percentage, "%)"
  ),
  motif_counts$Motif
)

# Plot 
pie_plot <- ggplot(motif_counts,
                   aes(x = "", y = Count, fill = fct_inorder(Motif))) +

  geom_col(width = 1, color = "white", linewidth = 0.7) +
  coord_polar(theta = "y", start = 0) +
  scale_fill_manual(values = motif_colors, labels = legend_labels) +

  # External labels with automatic arrows — without overlap
  geom_label_repel(
    data          = motif_pos,
    aes(y = pos, label = paste0(Motif, "\n", Percentage, "%")),
    size          = 3.2,
    color         = "grey10",
    fill          = "white",
    label.size    = 0.25,
    label.padding = unit(0.18, "lines"),
    segment.color = "grey10",
    segment.size  = 0.4,
    nudge_x       = 0.55,
    show.legend   = FALSE,
    max.overlaps  = Inf,
    force         = 12,
    family        = "sans"
  ) +

  theme_void(base_size = 11) +
  theme(
    legend.position      = "right",
    legend.title         = element_text(size = 10, face = "bold",
                                        color = "grey10", family = "mono"),
    legend.text          = element_text(size = 8.5, color = "grey10",
                                        family = "mono"),
    legend.key.size      = unit(0.42, "cm"),
    legend.spacing.y     = unit(0.15, "cm"),

    plot.title           = element_text(size = 13, face = "bold", hjust = 0.5,
                                        color = "grey10", margin = margin(b = 4)),
    plot.subtitle        = element_text(size = 9, hjust = 0.5, color = "grey10",
                                        margin = margin(b = 4)),
    plot.caption         = element_text(size = 7.5, color = "grey10",
                                        hjust = 0.5, margin = margin(t = 10)),
    plot.margin          = margin(14, 40, 14, 40),  # margem lateral generosa
    plot.background      = element_rect(fill = "white", color = NA),
    panel.background     = element_rect(fill = "white", color = NA)
  ) +

  labs(
    title    = "Distribution of Discovered SSR loci by Motif Types",
    subtitle = "Simple sequence repeats classified by repeat unit length",
    fill     = "Motif type",
    caption  = paste0("Total SSRs: ",
                      format(sum(motif_counts$Count), big.mark = ","))
  )

# Export
ggsave("motif_distribution_publication.png", pie_plot,
       width = 200, height = 160, units = "mm", dpi = 600, bg = "white")

ggsave("motif_distribution_publication.tiff", pie_plot,
       width = 200, height = 160, units = "mm", dpi = 600,
       compression = "lzw", bg = "white")
```

# 12. Scatter plot: % of genome in SSRs per chromosome

```{r}
# Processing data 
scatter_data <- final_table %>%
  mutate(Type = "Chromosome")

genome_total <- data.frame(
  Chromosome     = "Genome Total",
  Pct_chr_in_SSRs = (sum(final_table$Total_SSR_length_bp) / TOTAL_CHR_SIZE_BP) * 100,
  Type           = "Genome Total"
)

scatter_combined <- bind_rows(
  scatter_data %>% dplyr::select(Chromosome, Pct_chr_in_SSRs, Type),
  genome_total
)

scatter_combined$Chromosome <- factor(
  scatter_combined$Chromosome,
  levels = c(sort(unique(final_table$Chromosome)), "Genome Total")
)

# Valor da linha de referência
genome_pct <- genome_total$Pct_chr_in_SSRs

# Plot
scatter_plot <- ggplot(scatter_combined,
                       aes(x = Chromosome, y = Pct_chr_in_SSRs,
                           color = Type, shape = Type)) +

  # Subtle vertical grid line to guide the eye across chromosomes
  geom_vline(
    xintercept = seq_along(levels(scatter_combined$Chromosome)),
    color = "grey92", linewidth = 0.4
  ) +

  # Genome total reference line — behind the points
  geom_hline(
    yintercept = genome_pct,
    linetype   = "dashed",
    color      = "#CC79A7",
    linewidth  = 0.6,
    alpha      = 0.8
  ) +

  # Lollipop: stem from the point to y = 0 — adds visual weight without clutter
  geom_segment(
    data = filter(scatter_combined, Type == "Chromosome"),
    aes(x = Chromosome, xend = Chromosome,
        y = 0, yend = Pct_chr_in_SSRs),
    color     = "#56B4E9",
    linewidth = 0.5,
    alpha     = 0.5
  ) +

  # Points
  geom_point(size = 3.8, alpha = 0.95) +

  # Okabe-Ito palette — mesma do pie chart
  scale_color_manual(
    values = c("Chromosome" = "#0072B2", "Genome Total" = "#CC79A7")
  ) +
  scale_shape_manual(
    values = c("Chromosome" = 19, "Genome Total" = 18)
  ) +

  # Y-axis starts at 0
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.08)),
    labels = function(x) paste0(x, "%")
  ) +

  # Reference line annotation (avoids redundant legend)
  annotate(
    "text",
    x     = length(levels(scatter_combined$Chromosome)) - 0.3,
    y     = genome_pct + 0.05,
    label = paste0("Genome avg: ", round(genome_pct, 2), "%"),
    color = "#CC79A7",
    size  = 3.2,
    hjust = 1,
    fontface = "italic"
  ) +

  theme_bw(base_size = 11) +
  theme(
    # Axes
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 9,
                                    color = "grey20"),
    axis.text.y      = element_text(size = 9, color = "grey20"),
    axis.title       = element_text(size = 11, face = "bold", color = "grey10"),
    axis.ticks       = element_line(color = "grey80", linewidth = 0.35),

    # Grid — horizontal only
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.4),
    panel.border       = element_rect(color = "grey80", linewidth = 0.5),

    # Legend
    legend.position    = "top",
    legend.title       = element_blank(),
    legend.text        = element_text(size = 10, color = "grey20"),
    legend.key.size    = unit(0.45, "cm"),
    legend.spacing.x   = unit(0.3, "cm"),

    # Titles
    plot.title         = element_text(size = 13, face = "bold", hjust = 0.5,
                                      color = "grey10", margin = margin(b = 4)),
    plot.subtitle      = element_text(size = 9, hjust = 0.5, color = "grey50",
                                      margin = margin(b = 8)),
    plot.caption       = element_text(size = 7.5, color = "grey55",
                                      hjust = 0.5, margin = margin(t = 8)),

    # Background
    plot.background    = element_rect(fill = "white", color = NA),
    panel.background   = element_rect(fill = "white", color = NA),
    plot.margin        = margin(14, 20, 14, 14)
  ) +

  labs(
    title    = "Percentage of Genome Occupied by SSRs per Chromosome",
    subtitle = paste0("Genome total: ",
                      format(TOTAL_CHR_SIZE_BP, big.mark = ","), " bp"),
    x       = "Chromosome",
    y       = "% Chromosome in SSRs",
    caption = paste0("Genome average: ", round(genome_pct, 2), "%")
  )

# Export 
ggsave("genome_ssr_percentage_scatter.png", scatter_plot,
       width = 220, height = 140, units = "mm", dpi = 600, bg = "white")

ggsave("genome_ssr_percentage_scatter.tiff", scatter_plot,
       width = 220, height = 140, units = "mm", dpi = 600,
       compression = "lzw", bg = "white")
```

# 13. Boxplot: STR length by motif

```{r}
# Processing data 
boxplot_data <- str_data %>%
  filter(STR_type != "Other") %>%
  mutate(
    STR_type = factor(STR_type,
                      levels = c("Mono", "Di", "Tri", "Tetra", "Penta",
                                 "Hexa", "Hepta", "Octa", "Nona"))
  )

boxplot_stats <- boxplot_data %>%
  group_by(STR_type) %>%
  summarise(
    median_length = median(STR_length),
    mean_length   = mean(STR_length),
    n             = n(),
    .groups       = "drop"
  )

# Okabe-Ito palette 
motif_colors <- c(
  "Mono"  = "#56B4E9", "Di"    = "#0072B2", "Tri"   = "#009E73",
  "Tetra" = "#F0E442", "Penta" = "#E69F00", "Hexa"  = "#D55E00",
  "Hepta" = "#CC79A7", "Octa"  = "#999999", "Nona"  = "red"
)

# Versões mais escuras para a borda de cada box 
motif_colors_dark <- c(
  "Mono"  = "#2185C5", "Di"    = "#004E8C", "Tri"   = "#006B4E",
  "Tetra" = "#C4B800", "Penta" = "#B07800", "Hexa"  = "#8C3A00",
  "Hepta" = "#A0527F", "Octa"  = "#666666", "Nona"  = "darkred"
)

# Anotação n= por grupo 
n_labels <- boxplot_stats %>%
  mutate(label = paste0("n=", format(n, big.mark = ",")))

# Plot 
boxplot_plot <- ggplot(boxplot_data,
                       aes(x = STR_type, y = STR_length, fill = STR_type)) +

  # Main box
  geom_boxplot(
    aes(color = STR_type),
    outlier.shape  = 21,
    outlier.fill   = "white",
    outlier.color  = "grey60",
    outlier.alpha  = 0.35,
    outlier.size   = 0.8,
    linewidth      = 0.5,
    width          = 0.6,
    fatten         = 1.5    # mediana um pouco mais fina e elegante
  ) +

  # Median as text above the box
  stat_summary(
    fun      = median,
    geom     = "text",
    aes(label = sprintf("%.0f bp", after_stat(y))),
    vjust    = -0.6,
    size     = 3.0,
    color    = "grey25",
    fontface = "bold",
    family   = "sans"
  ) +

  # n= below each box (genomic figure standard)
  geom_text(
    data     = n_labels,
    aes(x = STR_type, y = -Inf, label = label),
    vjust    = -0.5,
    size     = 2.8,
    color    = "grey50",
    family   = "sans",
    inherit.aes = FALSE
  ) +

  scale_fill_manual(values = motif_colors) +
  scale_color_manual(values = motif_colors_dark) +

  scale_y_continuous(
  labels = scales::comma,
  expand = expansion(mult = c(0.05, 0.12))
) +

  theme_bw(base_size = 11) +
  theme(
    # Axes
    axis.text.x      = element_text(size = 10, color = "grey15", face = "bold"),
    axis.text.y      = element_text(size = 9,  color = "grey20"),
    axis.title       = element_text(size = 11, face = "bold", color = "grey10"),
    axis.ticks       = element_line(color = "grey80", linewidth = 0.35),

    # Grid — horizontal only
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "grey92", linewidth = 0.4),
    panel.border       = element_rect(color = "grey80", linewidth = 0.5),

    # No legend (color is already represented on the X-axis)
    legend.position  = "none",

    # Titles
    plot.title       = element_text(size = 13, face = "bold", hjust = 0.5,
                                    color = "grey10", margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 9,  hjust = 0.5, color = "grey50",
                                    margin = margin(b = 8)),
    plot.caption     = element_text(size = 7.5, color = "grey55",
                                    hjust = 0.5, margin = margin(t = 8)),

    # Background
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin      = margin(14, 20, 20, 14)
  ) +

  labs(
    title    = "Distribution of STR Length by Motif Type",
    subtitle = "Median values shown above each box",
    x        = "Motif type",
    y        = "STR length (bp)",
    caption  = paste0("Total STRs: ",
                      format(sum(boxplot_stats$n), big.mark = ","))
  )

# Export
ggsave("str_length_by_motif_boxplot.png", boxplot_plot,
       width = 220, height = 145, units = "mm", dpi = 600, bg = "white")

ggsave("str_length_by_motif_boxplot.tiff", boxplot_plot,
       width = 220, height = 145, units = "mm", dpi = 600,
       compression = "lzw", bg = "white")

cat("\n=== Analysis Complete ===\n")
cat("All plots have been saved.\n")
```

# 14. Physical map of motifs discovered via TRF (heatmap)

```{r}
# Processing data
# Derive chromosome sizes from the discovered regions
chrom_sizes_disc <- str_data %>%
  group_by(Chromosome) %>%
  summarise(size_bp = max(End), .groups = "drop") %>%
  rename(chr = Chromosome)

map_data_disc <- str_data %>%
  filter(STR_type != "Other") %>%
  mutate(
    chr          = Chromosome,
    position_mb  = Start / 1e6,
    motif_class_plot = factor(STR_type,
                              levels = c("Mono", "Di", "Tri", "Tetra",
                                         "Penta", "Hexa", "Hepta", "Octa", "Nona"))
  )

chrom_plot_data_disc <- chrom_sizes_disc %>%
  mutate(size_mb = size_bp / 1e6) %>%
  filter(chr %in% unique(map_data_disc$chr))

# Parameter: window size
window_mb <- 2

# Density bins
density_data_disc <- map_data_disc %>%
  mutate(bin = floor(position_mb / window_mb) * window_mb) %>%
  count(chr, bin, name = "n_loci") %>%
  left_join(chrom_plot_data_disc %>% select(chr, size_mb), by = "chr") %>%
  group_by(chr) %>%
  complete(bin = seq(0, max(size_mb), by = window_mb),
           fill = list(n_loci = 0)) %>%
  filter(bin < size_mb) %>%
  mutate(
    tile_top    = pmin(bin + window_mb, size_mb),
    tile_center = (bin + tile_top) / 2,
    tile_height = tile_top - bin
  ) %>%
  ungroup()

# Okabe-Ito palette
motif_colors <- c(
  "Mono"  = "#56B4E9", "Di"    = "#0072B2", "Tri"   = "#009E73",
  "Tetra" = "#F0E442", "Penta" = "#E69F00", "Hexa"  = "#D55E00",
  "Hepta" = "#CC79A7", "Octa"  = "#999999", "Nona"  = "#2C2C2A"
)

# Plot
physical_map_disc <- ggplot() +

  # Chromosome outline
  geom_rect(
    data = chrom_plot_data_disc,
    aes(xmin = as.numeric(factor(chr)) - 0.42,
        xmax = as.numeric(factor(chr)) + 0.42,
        ymin = 0, ymax = size_mb),
    fill = "grey97", color = "grey65", linewidth = 0.4
  ) +

  # Density heatmap
  geom_tile(
    data  = density_data_disc,
    aes(x = chr, y = tile_center, fill = n_loci, height = tile_height),
    width = 0.82
  ) +

  scale_fill_gradientn(
    colors = c("grey97", "#C6DBEF", "#6BAED6", "#2171B5", "#08306B"),
    values = scales::rescale(c(0, 0.01, 0.25, 0.6, 1)),
    name   = "SSR density\n(loci / bin)",
    guide  = guide_colorbar(
      barwidth     = 0.8,
      barheight    = 8,
      ticks        = TRUE,
      frame.colour = "grey60"
    )
  ) +

  scale_y_continuous(
    expand = expansion(mult = c(0, 0.03)),
    breaks = seq(0, 500, by = 50),
    labels = function(x) paste0(x, " Mb")
  ) +

  theme_bw(base_size = 11) +
  theme(
    axis.text.x      = element_text(size = 9, color = "grey20",
                                    angle = 45, hjust = 1),
    axis.text.y      = element_text(size = 9, color = "grey20"),
    axis.title       = element_text(size = 11, face = "bold", color = "grey10"),
    axis.ticks       = element_line(color = "grey80", linewidth = 0.35),

    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.35),
    panel.border       = element_rect(color = "grey75", linewidth = 0.5),

    legend.position    = "right",
    legend.title       = element_text(size = 10, face = "bold", color = "grey10"),
    legend.text        = element_text(size = 8.5, color = "grey25"),

    plot.title         = element_text(size = 13, face = "bold", hjust = 0.5,
                                      color = "grey10", margin = margin(b = 4)),
    plot.subtitle      = element_text(size = 9, hjust = 0.5, color = "grey50",
                                      margin = margin(b = 8)),
    plot.caption       = element_text(size = 7.5, color = "grey55",
                                      hjust = 0.5, margin = margin(t = 8)),
    plot.background    = element_rect(fill = "white", color = NA),
    panel.background   = element_rect(fill = "white", color = NA),
    plot.margin        = margin(14, 14, 14, 14)
  ) +

  labs(
    title    = "Heat Map: Genomic Distribution of Discovered SSRs",
    subtitle = paste0(format(nrow(map_data_disc), big.mark = ","),
                      " loci across ",
                      length(unique(map_data_disc$chr)),
                      " chromosomes  |  Window size: ", window_mb,
                      " Mb  |  Total: ",
                      format(TOTAL_CHR_SIZE_BP / 1e6, digits = 4), " Mb"),
    x       = "Chromosome",
    y       = "Position (Mb)",
    caption = paste0("Density calculated in ", window_mb,
                     " Mb non-overlapping windows")
  )

# Export
ggsave("physical_map_ssr_discovered_heat.png", physical_map_disc,
       width = 220, height = 180, units = "mm", dpi = 600, bg = "white")

ggsave("physical_map_ssr_discovered_heat.tiff", physical_map_disc,
       width = 220, height = 180, units = "mm", dpi = 600,
       compression = "lzw", bg = "white")
```

> Versão de dist. dos motif

```{r}
# Processing data
chrom_sizes_disc <- str_data %>%
  group_by(Chromosome) %>%
  summarise(size_bp = max(End), .groups = "drop") %>%
  rename(chr = Chromosome)

map_data_disc <- str_data %>%
  filter(STR_type != "Other") %>%
  mutate(
    chr              = Chromosome,
    position_mb      = Start / 1e6,
    motif_class_plot = factor(STR_type,
                              levels = c("Mono", "Di", "Tri", "Tetra",
                                         "Penta", "Hexa", "Hepta", "Octa", "Nona"))
  )

chrom_plot_data_disc <- chrom_sizes_disc %>%
  mutate(size_mb = size_bp / 1e6) %>%
  filter(chr %in% unique(map_data_disc$chr))

# Okabe-Ito palette
motif_colors <- c(
  "Mono"  = "#56B4E9", "Di"    = "#0072B2", "Tri"   = "#009E73",
  "Tetra" = "#F0E442", "Penta" = "#E69F00", "Hexa"  = "#D55E00",
  "Hepta" = "#CC79A7", "Octa"  = "#999999", "Nona"  = "#2C2C2A"
)

# Legend com n por motif
motif_n <- map_data_disc %>%
  count(motif_class_plot, name = "n") %>%
  mutate(label = paste0(motif_class_plot, "  (n=", format(n, big.mark = ","), ")"))

legend_labels <- setNames(motif_n$label, motif_n$motif_class_plot)

# Plot
physical_map_disc <- ggplot() +

  # Chromosome outline
  geom_rect(
    data = chrom_plot_data_disc,
    aes(xmin = as.numeric(factor(chr)) - 0.38,
        xmax = as.numeric(factor(chr)) + 0.38,
        ymin = 0, ymax = size_mb),
    fill      = "grey96",
    color     = "grey70",
    linewidth = 0.4
  ) +

  # Locus marks by motif
  geom_linerange(
    data = map_data_disc,
    aes(x     = chr,
        ymin  = position_mb,
        ymax  = position_mb + 0.08,
        color = motif_class_plot),
    linewidth = 18.0,
    alpha     = 10.0
  ) +

  scale_color_manual(values = motif_colors, labels = legend_labels,
                     name = "Motif type") +

  scale_y_continuous(
    expand = expansion(mult = c(0, 0.03)),
    breaks = seq(0, 500, by = 50),
    labels = function(x) paste0(x, " Mb")
  ) +

  theme_bw(base_size = 11) +
  theme(
    axis.text.x        = element_text(size = 9, color = "grey20",
                                      angle = 45, hjust = 1),
    axis.text.y        = element_text(size = 9, color = "grey20"),
    axis.title         = element_text(size = 11, face = "bold", color = "grey10"),
    axis.ticks         = element_line(color = "grey80", linewidth = 0.35),

    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.35),
    panel.border       = element_rect(color = "grey75", linewidth = 0.5),

    legend.position    = "right",
    legend.title       = element_text(size = 10, face = "bold", color = "grey10"),
    legend.text        = element_text(size = 8.5, color = "grey25"),
    legend.key.size    = unit(0.4, "cm"),
    legend.spacing.y   = unit(0.15, "cm"),

    plot.title         = element_text(size = 13, face = "bold", hjust = 0.5,
                                      color = "grey10", margin = margin(b = 4)),
    plot.subtitle      = element_text(size = 9, hjust = 0.5, color = "grey50",
                                      margin = margin(b = 8)),
    plot.caption       = element_text(size = 7.5, color = "grey55",
                                      hjust = 0.5, margin = margin(t = 8)),
    plot.background    = element_rect(fill = "white", color = NA),
    panel.background   = element_rect(fill = "white", color = NA),
    plot.margin        = margin(14, 14, 14, 14)
  ) +

  labs(
    title    = "Physical Map: Genomic Distribution of Discovered SSRs",
    subtitle = paste0(format(nrow(map_data_disc), big.mark = ","),
                      " loci across ",
                      length(unique(map_data_disc$chr)),
                      " chromosomes  |  Total: ",
                      format(TOTAL_CHR_SIZE_BP / 1e6, digits = 4), " Mb"),
    x       = "Chromosome",
    y       = "Position (Mb)",
    caption = "Each mark represents one SSR locus"
  )

# Export
ggsave("physical_map_ssr_discovered.png", physical_map_disc,
       width = 220, height = 180, units = "mm", dpi = 600, bg = "white")

ggsave("physical_map_ssr_discovered.tiff", physical_map_disc,
       width = 220, height = 180, units = "mm", dpi = 600,
       compression = "lzw", bg = "white")
```

