---
title: "Descriptive Analysis of STR Loci — HipSTR VCF"
author: "Rodrigues-Silva A."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.align = "center"
)
```

---

# NOTE ON USE

This script is compatible with two datasets produced in successive stages
of the HipSTR pipeline:

- **Dataset 2** — VCF after `filter_vcf.py` with genotype quality filters
  genótipo (`--min-call-qual 0.9`, `--max-call-flank-indel 0.15`,
  `--max-call-stutter 0.15`, `--min-call-allele-bias -2`,
  Genotypes that failed are **removed**
  (GT field = `./.`); there is no genotype-level `FILTER` tag in the FORMAT field.
  Input file: `hipstr_trf_v2_filtered.vcf`.

- **Dataset 3** — VCF with additional DP ≥ 8 and call rate ≥ 0.7 filters
  applied via DumpSTR, reducing the dataset from 29,375 to 1,999 loci.
  Valid genotypes carry `FILTER = PASS` in the FORMAT field.
  Input file: `hipstr_post_dumpstr_filtered_trf.vcf`.

Run only the sections marked with the dataset corresponding to the loaded file.
Unmarked sections are compatible with both.

---

# 1. Packages

```{r pacotes}
library(vcfR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(stringr)
library(readr)
library(ggridges)
library(patchwork)
library(scales)
library(ggrepel)
library(forcats)
```

---

# 2. Publication theme and color palette

```{r tema}
tema_pub <- theme_classic(base_size = 11, base_family = "sans") +
  theme(
    plot.title        = element_text(size = 13, face = "bold", hjust = 0,
                                     margin = margin(b = 6)),
    plot.subtitle     = element_text(size = 9.5, color = "grey40", hjust = 0,
                                     margin = margin(b = 10)),
    axis.title        = element_text(size = 10.5, face = "bold"),
    axis.text         = element_text(size = 9, color = "grey20"),
    axis.line         = element_line(color = "grey30", linewidth = 0.4),
    axis.ticks        = element_line(color = "grey30", linewidth = 0.4),
    legend.title      = element_text(size = 9.5, face = "bold"),
    legend.text       = element_text(size = 8.5),
    legend.key.size   = unit(0.45, "cm"),
    legend.background = element_blank(),
    legend.position   = "right",
    panel.background  = element_blank(),
    panel.grid.major  = element_line(color = "grey92", linewidth = 0.35),
    panel.grid.minor  = element_blank(),
    strip.background  = element_rect(fill = "grey93", color = NA),
    strip.text        = element_text(size = 9, face = "bold"),
    plot.margin       = margin(12, 14, 10, 12)
  )

cores_okabe <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#999999", "#000000"
)

theme_set(tema_pub)
```

---

# DATASET 2
# File: hipstr_trf_v2_filtered.vcf

## D2-3. VCF reading

```{r d2_leitura}
vcf <- read.vcfR("hipstr_trf_v2_filtered.vcf", verbose = FALSE)
geno_test <- extract.info(vcf, element = "START", as.numeric = FALSE, mask = FALSE)

samples    <- colnames(vcf@gt)[-1]
n_amostras <- length(samples)
n_locos    <- nrow(vcf@gt)

loco_ids <- vcf@fix[, "ID"]
if (all(loco_ids == "."))
  loco_ids <- paste0(vcf@fix[, "CHROM"], ":", vcf@fix[, "POS"])

cat("Genotyped loci :", n_locos, "\n")
cat("Samples          :", n_amostras, "\n")
cat("Samples          :", paste(samples, collapse = ", "), "\n")
```

## D2-4. Extraction of FORMAT matrices

```{r d2_matrizes_format}
dp_mat     <- extract.gt(vcf, "DP",          as.numeric = TRUE)
q_mat      <- extract.gt(vcf, "Q",           as.numeric = TRUE)
pq_mat     <- extract.gt(vcf, "PQ",          as.numeric = TRUE)
gldiff_mat <- extract.gt(vcf, "GLDIFF",      as.numeric = TRUE)
dsnp_mat   <- extract.gt(vcf, "DSNP",        as.numeric = TRUE)
dstut_mat  <- extract.gt(vcf, "DSTUTTER",    as.numeric = TRUE)
dflank_mat <- extract.gt(vcf, "DFLANKINDEL", as.numeric = TRUE)
ab_mat     <- extract.gt(vcf, "AB",          as.numeric = TRUE)
fs_mat     <- extract.gt(vcf, "FS",          as.numeric = TRUE)
dab_mat    <- extract.gt(vcf, "DAB",         as.numeric = TRUE)
gb_mat     <- extract.gt(vcf, "GB",          as.numeric = FALSE)
gt_mat     <- extract.gt(vcf, "GT",          as.numeric = FALSE)
```

## D2-5. Extraction of INFO fields

```{r d2_info_df}
info_df <- as.data.frame(vcf@fix) %>%
  mutate(
    PERIOD = as.integer(str_extract(INFO, "(?<=PERIOD=)[0-9]+")),
    HET    = as.numeric(str_extract(INFO, "(?<=HET=)[0-9.]+")),
    HWEP   = as.numeric(str_extract(INFO, "(?<=HWEP=)[0-9.e+-]+")),
    AN     = as.integer(str_extract(INFO, "(?<=AN=)[0-9]+")),
    NFILT  = as.integer(str_extract(INFO, "(?<=NFILT=)[0-9]+")),
    NSKIP  = as.integer(str_extract(INFO, "(?<=NSKIP=)[0-9]+")),
    Loco   = loco_ids
  )
```

## D2-6. Derived data

```{r d2_dados_derivados}
# Logical mask: TRUE = called genotype (GT present and not "./.")
gt_called <- gt_mat != "./." & !is.na(gt_mat)

# Call rate per sample and per locus
callrate_amostra <- tibble(
  Amostra      = samples,
  n_chamado    = colSums(gt_called, na.rm = TRUE),
  n_total      = nrow(gt_called),
  callrate_pct = round(colMeans(gt_called, na.rm = TRUE) * 100, 2)
)

callrate_loco <- tibble(
  Loco         = loco_ids,
  PERIOD       = info_df$PERIOD,
  n_chamado    = rowSums(gt_called, na.rm = TRUE),
  n_total      = ncol(gt_called),
  callrate_pct = round(rowMeans(gt_called, na.rm = TRUE) * 100, 2)
)

# DP of called genotypes (GT present)
dp_called <- as.data.frame(dp_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "DP") %>%
  left_join(
    as.data.frame(gt_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "GT"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(GT != "./.", !is.na(GT), !is.na(DP))

# GB (bp diff vs ref) of called genotypes
gb_long <- as.data.frame(gb_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "GB") %>%
  left_join(
    as.data.frame(gt_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "GT"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(GT != "./.", !is.na(GT), !is.na(GB), GB != ".") %>%
  mutate(
    al1 = suppressWarnings(as.integer(str_extract(GB, "^-?[0-9]+"))),
    al2 = suppressWarnings(as.integer(str_extract(GB, "(?<=\\|)-?[0-9]+")))
  ) %>%
  pivot_longer(c(al1, al2), values_to = "bp_diff") %>%
  filter(!is.na(bp_diff)) %>%
  left_join(info_df %>% select(Loco, PERIOD), by = "Loco")

# All DP observations (including low-DP genotypes, without filtering)
dp_all <- as.data.frame(dp_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "DP") %>%
  filter(!is.na(DP))
```

## D2-7. Global descriptive statistics

```{r d2_stats_globais}
resumo_campo <- function(mat, nome) {
  vals <- as.numeric(mat[!is.na(mat)])
  tibble(
    Campo   = nome,
    N_obs   = length(vals),
    N_NA    = sum(is.na(mat)),
    Min     = min(vals,  na.rm = TRUE),
    Q25     = quantile(vals, 0.25, na.rm = TRUE),
    Mediana = median(vals, na.rm = TRUE),
    Media   = round(mean(vals, na.rm = TRUE), 4),
    Q75     = quantile(vals, 0.75, na.rm = TRUE),
    Max     = max(vals,  na.rm = TRUE),
    SD      = round(sd(vals, na.rm = TRUE), 4)
  )
}

tabela_global <- bind_rows(
  resumo_campo(dp_mat,     "DP"),
  resumo_campo(q_mat,      "Q"),
  resumo_campo(pq_mat,     "PQ"),
  resumo_campo(gldiff_mat, "GLDIFF"),
  resumo_campo(dsnp_mat,   "DSNP"),
  resumo_campo(dstut_mat,  "DSTUTTER"),
  resumo_campo(dflank_mat, "DFLANKINDEL"),
  resumo_campo(ab_mat,     "AB"),
  resumo_campo(fs_mat,     "FS"),
  resumo_campo(dab_mat,    "DAB")
)

knitr::kable(tabela_global, digits = 4)
write.csv(tabela_global, "tabela_global_D2.csv", row.names = FALSE)
```

## D2-8. Statistics per locus

```{r d2_stats_por_loco}
stats_por_loco <- tibble(
  Loco             = loco_ids,
  CHROM            = vcf@fix[, "CHROM"],
  POS              = as.integer(vcf@fix[, "POS"]),
  DP_mean          = rowMeans(dp_mat,     na.rm = TRUE),
  DP_min           = apply(dp_mat, 1, min, na.rm = TRUE),
  DP_max           = apply(dp_mat, 1, max, na.rm = TRUE),
  Q_mean           = rowMeans(q_mat,      na.rm = TRUE),
  Q_min            = apply(q_mat, 1, min, na.rm = TRUE),
  PQ_mean          = rowMeans(pq_mat,     na.rm = TRUE),
  GLDIFF_mean      = rowMeans(gldiff_mat, na.rm = TRUE),
  GLDIFF_min       = apply(gldiff_mat, 1, min, na.rm = TRUE),
  DSTUTTER_mean    = rowMeans(dstut_mat,  na.rm = TRUE),
  DSTUTTER_max     = apply(dstut_mat, 1, max, na.rm = TRUE),
  DFLANKINDEL_mean = rowMeans(dflank_mat, na.rm = TRUE),
  AB_mean          = rowMeans(ab_mat,     na.rm = TRUE),
  AB_min           = apply(ab_mat, 1, min, na.rm = TRUE),
  FS_mean          = rowMeans(fs_mat,     na.rm = TRUE),
  FS_min           = apply(fs_mat, 1, min, na.rm = TRUE),
  N_chamado        = rowSums(gt_called,  na.rm = TRUE),
  N_missing        = rowSums(!gt_called, na.rm = TRUE),
  Callrate_pct     = round(rowMeans(gt_called, na.rm = TRUE) * 100, 1)
) %>%
  mutate(across(where(is.double), ~ round(.x, 4)))

cat("Call rate distribution per locus:\n")
summary(stats_por_loco$Callrate_pct)

knitr::kable(head(stats_por_loco, 10))
```

## D2-9. Statistics per sample

```{r d2_stats_por_amostra}
stats_por_amostra <- tibble(
  Amostra       = samples,
  DP_mean       = colMeans(dp_mat,     na.rm = TRUE),
  DP_min        = apply(dp_mat, 2, min, na.rm = TRUE),
  DP_max        = apply(dp_mat, 2, max, na.rm = TRUE),
  Q_mean        = colMeans(q_mat,      na.rm = TRUE),
  Q_min         = apply(q_mat, 2, min, na.rm = TRUE),
  GLDIFF_mean   = colMeans(gldiff_mat, na.rm = TRUE),
  DSTUTTER_mean = colMeans(dstut_mat,  na.rm = TRUE),
  AB_mean       = colMeans(ab_mat,     na.rm = TRUE),
  FS_mean       = colMeans(fs_mat,     na.rm = TRUE),
  N_chamado     = colSums(gt_called,  na.rm = TRUE),
  N_missing     = colSums(!gt_called, na.rm = TRUE),
  Callrate_pct  = round(colMeans(gt_called, na.rm = TRUE) * 100, 1)
) %>%
  mutate(across(where(is.double), ~ round(.x, 4))) %>%
  arrange(desc(Callrate_pct))

knitr::kable(stats_por_amostra)
```

## D2-10. Raw call rate per locus

```{r d2_callrate_loco}
cat("Call rate distribution per locus:\n")
summary(callrate_loco$callrate_pct)

callrate_loco %>%
  count(
    faixa = cut(callrate_pct,
                breaks = c(0, 25, 50, 70, 90, 100),
                include.lowest = TRUE)
  ) %>%
  mutate(pct = round(n / sum(n) * 100, 2)) %>%
  knitr::kable(col.names = c("Call rate range (%)", "N loci", "% loci"))
```

## D2-11. GB variation (bp diff vs. reference)

```{r d2_gb_stats}
gb_nums_d2 <- gb_long$bp_diff

tibble(
  Estatistica = c("Min", "Q25", "Median", "Mean", "Q75", "Max", "SD"),
  Valor_bp    = c(
    min(gb_nums_d2),
    quantile(gb_nums_d2, 0.25),
    median(gb_nums_d2),
    round(mean(gb_nums_d2), 2),
    quantile(gb_nums_d2, 0.75),
    max(gb_nums_d2),
    round(sd(gb_nums_d2), 2)
  )
) %>% knitr::kable()
```

## D2-12. CSV table export

```{r d2_exportar_csv}
write_csv(tabela_global,     "d2_stats_global_campos_FORMAT.csv")
write_csv(stats_por_loco,    "d2_stats_por_loco.csv")
write_csv(stats_por_amostra, "d2_stats_por_amostra.csv")
write_csv(callrate_loco,     "d2_callrate_por_loco.csv")

cat("Files generated:\n",
    "  d2_stats_global_campos_FORMAT.csv\n",
    "  d2_stats_por_loco.csv\n",
    "  d2_stats_por_amostra.csv\n",
    "  d2_callrate_por_loco.csv\n")
```

## D2-13. Figure 1 — DP and call rate composition by sample

```{r d2_fig1, fig.width=10, fig.height=10}
# Depth categories
dp_cat <- as.data.frame(dp_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "DP") %>%
  filter(!is.na(DP)) %>%
  mutate(
    DP_cat = cut(DP,
                 breaks = c(0, 5, 10, 15, 20, 25, 30, Inf),
                 labels = c("0–5", "5–10", "10–15",
                            "15–20", "20–25", "25–30", "30+"),
                 right  = FALSE,
                 include.lowest = TRUE)
  ) %>%
  count(Amostra, DP_cat) %>%
  group_by(Amostra) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

cores_dp <- c(
  "0–5"   = "#D55E00",
  "5–10"  = "#E69F00",
  "10–15" = "#F0E442",
  "15–20" = "#009E73",
  "20–25" = "#56B4E9",
  "25–30" = "#0072B2",
  "30+"   = "#000040"
)

# Sort samples by cumulative proportion of DP ≥ 10 (decreasing)
ordem_amostras <- dp_cat %>%
  filter(DP_cat %in% c("10–15", "15–20", "20–25", "25–30", "30+")) %>%
  group_by(Amostra) %>%
  summarise(pct_alta = sum(pct)) %>%
  arrange(desc(pct_alta)) %>%
  pull(Amostra)

p1a <- dp_cat %>%
  mutate(Amostra = factor(Amostra, levels = ordem_amostras)) %>%
  ggplot(aes(x = Amostra, y = pct, fill = DP_cat)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = cores_dp, name = "DP (reads)") +
  scale_y_continuous(expand = c(0, 0),
                     labels = label_percent(scale = 1)) +
  labs(
    title    = "A  Read depth composition by sample",
    subtitle = "Proportion of loci by DP category; samples sorted by the proportion with DP ≥ 10",
    x        = "Sample",
    y        = "% of loci"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7.5))

p1b <- callrate_amostra %>%
  arrange(callrate_pct) %>%
  mutate(Amostra = factor(Amostra, levels = Amostra)) %>%
  ggplot(aes(x = callrate_pct, y = Amostra)) +
  geom_segment(aes(xend = 0, yend = Amostra),
               color = "grey70", linewidth = 0.5) +
  geom_point(aes(color = callrate_pct), size = 3.5) +
  scale_color_gradient2(
    low = "#D55E00", mid = "#E69F00", high = "#009E73",
    midpoint = 70, guide = "none"
  ) +
  geom_vline(xintercept = 70, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  scale_x_continuous(labels = label_percent(scale = 1),
                     limits = c(0, 100)) +
  labs(
    title    = "B  Call rate by sample",
    subtitle = "Proportion of loci with a called GT; dashed line = 70%",
    x        = "% of called loci",
    y        = NULL
  ) +
  theme(axis.text.y = element_text(size = 7.5))

fig1_d2 <- p1a / p1b + plot_layout(heights = c(1.3, 1))
fig1_d2

# Sample ordering in p1a uses the cumulative fraction of DP ≥ 10 as the criterion, which places the best-sequenced samples on the left and makes it immediately visible which have an excess of low-coverage loci.
# The "0–5" category threshold includes loci with DP = 0 that still have a called GT — this can occur if HipSTR assigned a genotype with support from flanking reads but without central reads.
```

## D2-14. Figure 2 — DP of called genotypes by sample

```{r d2_fig2, fig.width=10, fig.height=10}
p2a <- dp_called %>%
  ggplot(aes(x = DP)) +
  geom_histogram(binwidth = 1, fill = "#0072B2", color = "white",
                 linewidth = 0.15, alpha = 0.9) +
  geom_vline(xintercept = median(dp_called$DP), linetype = "dashed",
             color = "#D55E00", linewidth = 0.7) +
  annotate("text", x = median(dp_called$DP) + 1.5, y = Inf,
           label = paste0("Median = ", median(dp_called$DP)),
           hjust = 0, vjust = 1.5, size = 3, color = "#D55E00") +
  scale_x_continuous(breaks = c(8, 10, 15, 20, 30, 40, 50, 64)) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  labs(
    title    = "A  Distribution of DP (called genotypes)",
    subtitle = "Genotypes with GT present after running filter_vcf.py",
    x        = "DP (reads per genotype)",
    y        = "Frequency"
  )

p2b <- dp_called %>%
  mutate(Amostra = reorder(Amostra, DP, FUN = median)) %>%
  ggplot(aes(x = Amostra, y = DP)) +
  geom_boxplot(fill = "#56B4E9", color = "grey30", outlier.size = 0.6,
               outlier.alpha = 0.4, linewidth = 0.45, width = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 18,
               size = 2.2, color = "#D55E00") +
  scale_y_continuous(breaks = c(8, 10, 15, 20, 30, 40, 50, 64)) +
  labs(
    title    = "B  DP by sample (called genotypes)",
    subtitle = "Orange diamond = mean; boxes sorted by median",
    x        = "Sample",
    y        = "DP (reads)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7.5))

fig2_d2 <- p2a / p2b
fig2_d2
```

## D2-15. Figure 3 — Genotype quality (Q and GLDIFF)

```{r d2_fig3, fig.width=11, fig.height=10}
q_long <- as.data.frame(q_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "Q") %>%
  left_join(
    as.data.frame(gt_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "GT"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(GT != "./.", !is.na(GT), !is.na(Q))

gldiff_long <- as.data.frame(gldiff_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "GLDIFF") %>%
  left_join(
    as.data.frame(gt_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "GT"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(GT != "./.", !is.na(GT), !is.na(GLDIFF))

p3a <- q_long %>%
  ggplot(aes(x = Q)) +
  geom_histogram(binwidth = 0.005, boundary = 0.0025, fill = "#009E73", color = "white",
                 linewidth = 0.1, alpha = 0.9) +
  scale_x_continuous(limits = c(0.88, 1.005),
                     breaks = seq(0.88, 1.00, by = 0.02)) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  labs(
    title    = "A  Posterior probability of the genotype (Q)",
    subtitle = "Called genotypes; values concentrated close to 1.0",
    x        = "Q (posterior probability)",
    y        = "Frequency"
  )

p3b <- gldiff_long %>%
  ggplot(aes(x = GLDIFF)) +
  geom_histogram(bins = 60, fill = "#CC79A7", color = "white",
                 linewidth = 0.1, alpha = 0.9) +
  geom_vline(xintercept = 2, linetype = "dashed",
             color = "grey40", linewidth = 0.6) +
  annotate("text", x = 3, y = Inf, label = "GLDIFF = 2",
           hjust = 0, vjust = 1.5, size = 3, color = "grey30") +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100),
                labels = c(1, 2, 5, 10, 20, 50, 100)) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  labs(
    title    = "B  Likelihood difference between genotypes (GLDIFF)",
    subtitle = "Log10 scale; higher values indicate greater confidence in the call",
    x        = "GLDIFF (log10)",
    y        = "Frequency"
  )

set.seed(42)
qgl_joined <- q_long %>%
  inner_join(gldiff_long, by = c("Loco", "Amostra")) %>%
  sample_n(min(5000, n()))

p3c <- qgl_joined %>%
  ggplot(aes(x = GLDIFF, y = Q)) +
  geom_hex(bins = 50) +
  scale_fill_gradientn(
    colors = c("grey90", "#56B4E9", "#0072B2", "#000040"),
    name   = "N genotypes",
    labels = label_comma()
  ) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100)) +
  scale_y_continuous(limits = c(0.88, 1.005),
                     breaks = c(0.9, 0.95, 0.99, 1.0)) +
  labs(
    title    = "C  Q × GLDIFF (hexbin)",
    subtitle = "Relationship between posterior probability and relative confidence",
    x        = "GLDIFF (log10)",
    y        = "Q"
  )

fig3_d2 <- (p3a | p3b) / p3c
fig3_d2
```

## D2-16. Figure 4 — Artifacts: stutter, allele bias, and strand bias

```{r d2_fig6, fig.width=10, fig.height=10}
art_df <- tibble(
  Loco          = loco_ids,
  PERIOD        = info_df$PERIOD,
  DSTUTTER_mean = rowMeans(dstut_mat,  na.rm = TRUE),
  DFLANK_mean   = rowMeans(dflank_mat, na.rm = TRUE)
)

p6a <- art_df %>%
  filter(!is.na(PERIOD), PERIOD >= 2) %>%
  group_by(PERIOD) %>%
  summarise(
    pct_stutter = mean(DSTUTTER_mean > 0, na.rm = TRUE) * 100,
    n_locos     = n()
  ) %>%
  ggplot(aes(x = factor(PERIOD), y = pct_stutter, fill = factor(PERIOD))) +
  geom_col(color = "white", linewidth = 0.2, width = 0.7) +
  scale_fill_manual(values = cores_okabe[1:8], guide = "none") +
  scale_y_continuous(labels = label_percent(scale = 1), expand = c(0, 0)) +
  labs(
    title    = "A  % of loci with detected stutter (DSTUTTER > 0) by period",
    subtitle = "Stutter is an expected artifact in STRs, especially for short motifs",
    x        = "Period (bp)",
    y        = "% of loci"
  )

# AB and FS: filter by called GT (not "./.")
ab_long <- as.data.frame(ab_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "AB") %>%
  left_join(
    as.data.frame(gt_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "GT"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(GT != "./.", !is.na(GT), !is.na(AB), AB < 0)

fs_long <- as.data.frame(fs_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "FS") %>%
  left_join(
    as.data.frame(gt_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "GT"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(GT != "./.", !is.na(GT), !is.na(FS), FS < 0)

bias_long <- bind_rows(
  ab_long %>% transmute(valor = AB, Tipo = "AB (allele bias)"),
  fs_long %>% transmute(valor = FS, Tipo = "FS (strand bias)")
)

p6b <- bias_long %>%
  ggplot(aes(x = valor, fill = Tipo)) +
  geom_histogram(binwidth = 0.05, alpha = 0.8, color = "white",
                 linewidth = 0.1, position = "identity") +
  geom_vline(
  data = data.frame(x_int = -2, tipo_linha = "dashed"),
  aes(xintercept = x_int, linetype = I(tipo_linha)),
  color = "grey30", linewidth = 0.5
) +
  scale_fill_manual(values = c("#E69F00", "#CC79A7"), name = NULL) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  facet_wrap(~Tipo, scales = "free_y") +
  labs(
    title    = "B  Distribution of allele bias (AB) and strand bias (FS)",
    subtitle = "Called heterozygous genotypes; values = log10(p-value)",
    x        = "log10(bias p-value)",
    y        = "Frequency"
  ) +
  theme(legend.position = "none")

fig6_d2 <- p6a / p6b
fig6_d2
```
## D2-17. Figure 5 — Allelic variation (GB)

```{r d2_fig5, fig.width=10, fig.height=11}
p5a <- gb_long %>%
  ggplot(aes(x = bp_diff)) +
  geom_histogram(binwidth = 1, fill = "#0072B2", color = "white",
                 linewidth = 0.1, alpha = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#D55E00",
             linewidth = 0.7) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  labs(
    title    = "A  Distribution of allelic variation (GB)",
    subtitle = "Difference in bp of each called allele vs. reference allele",
    x        = "GB (bp diff. vs. reference)",
    y        = "Allele frequency"
  )

p5b <- gb_long %>%
  filter(PERIOD >= 2, PERIOD <= 9) %>%
  mutate(PERIOD_lab = paste0("PERIOD = ", PERIOD, " bp")) %>%
  ggplot(aes(x = bp_diff,
             y = fct_rev(factor(PERIOD_lab)),
             fill = factor(PERIOD))) +
  geom_density_ridges(
    alpha          = 0.75,
    color          = "grey40",
    linewidth      = 0.3,
    bandwidth      = 1.5,
    scale          = 1.1,
    rel_min_height = 0.01
  ) +
  scale_fill_manual(values = cores_okabe[1:8], guide = "none") +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey30", linewidth = 0.5) +
  labs(
    title    = "B  Allelic variation by repeat motif period (ridgeline)",
    subtitle = "Distribution of bp_diff conditioned on repeat period",
    x        = "GB (bp diff. vs. reference)",
    y        = NULL
  )

fig5_d2 <- p5a / p5b + plot_layout(heights = c(1, 1.6))
fig5_d2
```

## D2-18. Figure 6 — DP heatmap by sample × chromosome

```{r d2_fig7, fig.width=9, fig.height=8}
dp_chr <- as.data.frame(dp_mat) %>%
  mutate(Loco = loco_ids, CHROM = vcf@fix[, "CHROM"]) %>%
  filter(str_detect(CHROM, "^Dal_")) %>%
  pivot_longer(-c(Loco, CHROM), names_to = "Amostra", values_to = "DP") %>%
  left_join(
    as.data.frame(gt_mat) %>%
      mutate(Loco = loco_ids, CHROM = vcf@fix[, "CHROM"]) %>%
      pivot_longer(-c(Loco, CHROM), names_to = "Amostra", values_to = "GT"),
    by = c("Loco", "Amostra", "CHROM")
  ) %>%
  filter(GT != "./.", !is.na(GT)) %>%
  group_by(CHROM, Amostra) %>%
  summarise(DP_mean = mean(DP, na.rm = TRUE), .groups = "drop")

fig7_d2 <- dp_chr %>%
  mutate(
    CHROM   = factor(CHROM, levels = paste0("Dal_0", 1:8)),
    Amostra = reorder(Amostra, DP_mean)
  ) %>%
  ggplot(aes(x = CHROM, y = Amostra, fill = DP_mean)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradientn(
    colors = c("#f7fbff", "#c6dbef", "#6baed6", "#2171b5", "#08306b"),
    name   = "Mean DP"
  ) +
  geom_text(aes(label = round(DP_mean, 1)), size = 2.5, color = "white") +
  labs(
    title    = "Mean read depth (DP) by sample × chromosome",
    subtitle = "Called genotypes on main chromosomes (Dal_01–Dal_08)",
    x        = "Chromosome",
    y        = "Sample"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

fig7_d2
```
## D2-19. Physical map and locus density by motif type

```{r d2_fig_genomic_maps_motif, fig.width=16, fig.height=10}
# Load chromosome sizes
chrom_sizes <- read_tsv("chromosome_sizes.txt", 
                        col_names = c("chr", "size_bp"), 
                        show_col_types = FALSE)

total_chrom_size_bp <- sum(chrom_sizes$size_bp)

# DIAGNOSTICS
cat("=== DIAGNOSTICS D2 ===\n")
cat("Total loci in VCF:", n_locos, "\n")
cat("PERIOD extracted (non-NA):", sum(!is.na(info_df$PERIOD)), "\n")
cat("PERIOD with NA:", sum(is.na(info_df$PERIOD)), "\n\n")

# PERIOD sample
cat("Sample of extracted PERIOD values:\n")
print(head(info_df$PERIOD, 20))

# Prepare data with PERIOD - KEEP ALL LOCI
map_data <- tibble(
  locus = loco_ids,
  chr = vcf@fix[, "CHROM"],
  pos = as.numeric(vcf@fix[, "POS"]),
  position_mb = pos / 1e6,
  period = info_df$PERIOD
) %>%
  filter(!is.na(pos)) %>%  # Keep everything except invalid POS
  mutate(
    # Se PERIOD for NA, usar 0 como "unknown"
    period = ifelse(is.na(period), 0, period),
    motif_class = case_when(
      period == 0 ~ "Unknown",
      period == 1 ~ "Mono",
      period == 2 ~ "Di",
      period == 3 ~ "Tri",
      period == 4 ~ "Tetra",
      period == 5 ~ "Penta",
      period == 6 ~ "Hexa",
      period == 7 ~ "Hepta",
      period == 8 ~ "Octa",
      period == 9 ~ "Nona",
      TRUE ~ "Other"
    ),
    motif_class = factor(motif_class, 
                         levels = c("Mono", "Di", "Tri", "Tetra", "Penta", 
                                   "Hexa", "Hepta", "Octa", "Nona", "Unknown", "Other"))
  )

chrom_plot_data <- chrom_sizes %>%
  mutate(size_mb = size_bp / 1e6, chr = as.character(chr)) %>%
  filter(chr %in% unique(map_data$chr))

cat("Final loci on map:", nrow(map_data), "\n\n")

# Okabe-Ito palette for motifs (with Unknown)
motif_colors <- c(
  "Mono"  = "#56B4E9", "Di"    = "#0072B2", "Tri"   = "#009E73",
  "Tetra" = "#F0E442", "Penta" = "#E69F00", "Hexa"  = "#D55E00",
  "Hepta" = "#CC79A7", "Octa"  = "#999999", "Nona"  = "#2C2C2A",
  "Unknown" = "#DDDDDD", "Other" = "#CCCCCC"
)

# Legend with counts
motif_counts <- map_data %>%
  count(motif_class, name = "n") %>%
  mutate(
    pct = round((n / sum(n)) * 100, 1),
    label = paste0(motif_class, " (n=", format(n, big.mark = ","), 
                   ", ", pct, "%)")
  )

legend_labels <- setNames(motif_counts$label, motif_counts$motif_class)

cat("Distribution by motif:\n")
print(motif_counts)

# ---- PLOT 1: Physical Map by Motif ----
physical_map <- ggplot() +
  geom_rect(
    data = chrom_plot_data,
    aes(xmin = as.numeric(factor(chr)) - 0.38,
        xmax = as.numeric(factor(chr)) + 0.38,
        ymin = 0, ymax = size_mb),
    fill = "grey96", color = "grey70", linewidth = 0.4
  ) +
  geom_linerange(
    data = map_data,
    aes(x = chr, ymin = position_mb, ymax = position_mb + 0.08, 
        color = motif_class),
    linewidth = 16.0, alpha = 1.0
  ) +
  scale_color_manual(
    values = motif_colors, 
    labels = legend_labels,
    name = "Motif type",
    guide = guide_legend(ncol = 1)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.03)),
    breaks = seq(0, 700, by = 50),
    labels = function(x) paste0(x, " Mb")
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1, color = "grey20"),
    axis.text.y = element_text(size = 9, color = "grey20"),
    axis.title = element_text(size = 11, face = "bold", color = "grey10"),
    axis.line = element_line(color = "grey30", linewidth = 0.4),
    axis.ticks = element_line(color = "grey30", linewidth = 0.4),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.35),
    panel.border = element_rect(color = "grey75", linewidth = 0.5),
    legend.title = element_text(size = 9.5, face = "bold", color = "grey10"),
    legend.text = element_text(size = 8, color = "grey25"),
    legend.key.size = unit(0.35, "cm"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5,
                              color = "grey10", margin = margin(b = 6)),
    plot.subtitle = element_text(size = 9.5, color = "grey40", hjust = 0.5,
                                 margin = margin(b = 8)),
    plot.margin = margin(12, 12, 10, 12)
  ) +
  labs(
    title = "A  Physical map: STR loci by motif repeat period",
    subtitle = paste0(format(nrow(map_data), big.mark = ","), " loci across ",
                      length(unique(map_data$chr)), " chromosomes"),
    x = "Chromosome",
    y = "Position (Mb)"
  )

# ---- PLOT 2: Density Heatmap ----
window_mb <- 2.0

density_data <- map_data %>%
  mutate(bin = floor(position_mb / window_mb) * window_mb) %>%
  count(chr, bin, name = "n_loci") %>%
  left_join(chrom_plot_data %>% select(chr, size_mb), by = "chr") %>%
  group_by(chr) %>%
  do({
    bins_seq <- seq(0, floor(max(.$size_mb, na.rm = TRUE) / window_mb) * window_mb, 
                    by = window_mb)
    complete(., bin = bins_seq, fill = list(n_loci = 0))
  }) %>%
  filter(bin < size_mb) %>%
  mutate(
    tile_top = pmin(bin + window_mb, size_mb),
    tile_center = (bin + tile_top) / 2,
    tile_height = tile_top - bin
  ) %>%
  ungroup()

density_map <- ggplot() +
  geom_rect(
    data = chrom_plot_data,
    aes(xmin = as.numeric(factor(chr)) - 0.42,
        xmax = as.numeric(factor(chr)) + 0.42,
        ymin = 0, ymax = size_mb + 0.5),
    fill = "grey97", color = "grey65", linewidth = 0.4
  ) +
  geom_tile(
    data = density_data,
    aes(x = chr, y = tile_center, fill = n_loci, height = tile_height),
    width = 0.79, color = NA
  ) +
  scale_fill_gradientn(
    colors = c("white", "#B3D9FF", "#4D94FF", "#0052CC", "#001A4D"),
    name = "# loci",
    guide = guide_colorbar(barwidth = 0.7, barheight = 4, ticks = FALSE)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.03)),
    breaks = seq(0, 700, by = 50),
    labels = function(x) paste0(x, " Mb")
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1, color = "grey20"),
    axis.text.y = element_text(size = 9, color = "grey20"),
    axis.title = element_text(size = 11, face = "bold", color = "grey10"),
    axis.line = element_line(color = "grey30", linewidth = 0.4),
    axis.ticks = element_line(color = "grey30", linewidth = 0.4),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.35),
    panel.border = element_rect(color = "grey75", linewidth = 0.5),
    legend.title = element_text(size = 9.5, face = "bold", color = "grey10"),
    legend.text = element_text(size = 8, color = "grey25"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5,
                              color = "grey10", margin = margin(b = 6)),
    plot.subtitle = element_text(size = 9.5, color = "grey40", hjust = 0.5,
                                 margin = margin(b = 8)),
    plot.margin = margin(12, 12, 10, 12)
  ) +
  labs(
    title = "B  STR density heatmap across chromosomes",
    subtitle = paste0("Window size: ", window_mb, " Mb (PASS loci only)"),
    x = "Chromosome",
    y = "Position (Mb)"
  )

# Combine plots
fig_genomic_d2 <- physical_map | density_map
fig_genomic_d2

# Export
ggsave("d2_physical_map_by_motif_v2.png", plot = physical_map,
       width = 220, height = 160, units = "mm", dpi = 600, bg = "white")
ggsave("d2_density_heatmap_v2.png", plot = density_map,
       width = 220, height = 160, units = "mm", dpi = 600, bg = "white")

cat("\n✓ Figures exported:\n",
    "  d2_physical_map_by_motif.png\n",
    "  d2_density_heatmap.png\n")
``` 

---

# DATASET 3
# File: hipstr_post_dumpstr_filtered_trf.vcf

## D3-3. VCF reading

```{r d3_leitura}
vcf <- read.vcfR("hipstr_post_dumpstr_filtered_trf.vcf", verbose = FALSE)

samples    <- colnames(vcf@gt)[-1]
n_amostras <- length(samples)
n_locos    <- nrow(vcf@gt)

loco_ids <- vcf@fix[, "ID"]
if (all(loco_ids == "."))
  loco_ids <- paste0(vcf@fix[, "CHROM"], ":", vcf@fix[, "POS"])

cat("Locos genotipados :", n_locos, "\n")
cat("Amostras          :", n_amostras, "\n")
cat("Amostras          :", paste(samples, collapse = ", "), "\n")
```

## D3-4. Extraction of FORMAT matrices

```{r d3_matrizes_format}
dp_mat     <- extract.gt(vcf, "DP", as.numeric = TRUE)
q_mat      <- extract.gt(vcf, "Q", as.numeric = TRUE)
pq_mat     <- extract.gt(vcf, "PQ", as.numeric = TRUE)
gldiff_mat <- extract.gt(vcf, "GLDIFF", as.numeric = TRUE)
dsnp_mat   <- extract.gt(vcf, "DSNP", as.numeric = TRUE)
dstut_mat  <- extract.gt(vcf, "DSTUTTER", as.numeric = TRUE)
dflank_mat <- extract.gt(vcf, "DFLANKINDEL", as.numeric = TRUE)
ab_mat     <- extract.gt(vcf, "AB", as.numeric = TRUE)
fs_mat     <- extract.gt(vcf, "FS", as.numeric = TRUE)
dab_mat    <- extract.gt(vcf, "DAB", as.numeric = TRUE)
filter_mat <- extract.gt(vcf, "FILTER", as.numeric = FALSE)
gb_mat     <- extract.gt(vcf, "GB", as.numeric = FALSE)
gt_mat     <- extract.gt(vcf, "GT", as.numeric = FALSE)
het_mat    <- extract.info(vcf, "HET", as.numeric = FALSE)
hwep_mat   <- extract.info(vcf, "HWEP", as.numeric = FALSE)
```

## D3-5. Extraction of INFO fields

```{r d3_info_df}
info_df <- as.data.frame(vcf@fix) %>%
  mutate(
    PERIOD = as.integer(str_extract(INFO, "(?<=PERIOD=)[0-9]+")),
    HET    = as.numeric(str_extract(INFO, "(?<=HET=)[0-9.]+")),
    HWEP   = as.numeric(str_extract(INFO, "(?<=HWEP=)[0-9.e+-]+")),
    AN     = as.integer(str_extract(INFO, "(?<=AN=)[0-9]+")),
    NFILT  = as.integer(str_extract(INFO, "(?<=NFILT=)[0-9]+")),
    NSKIP  = as.integer(str_extract(INFO, "(?<=NSKIP=)[0-9]+")),
    Loco   = loco_ids
  )
```

## D3-6. Derived data

```{r d3_dados_derivados}
# % PASS per sample
perc_pass_amostra <- tibble(
  Amostra   = samples,
  perc_pass = apply(filter_mat, 2,
                    function(x) mean(x == "PASS", na.rm = TRUE) * 100)
)

# % PASS per locus
perc_pass_loco <- tibble(
  Loco      = loco_ids,
  perc_pass = apply(filter_mat, 1,
                    function(x) mean(x == "PASS", na.rm = TRUE) * 100),
  PERIOD    = info_df$PERIOD
)

# DP for PASS only
dp_pass <- as.data.frame(dp_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "DP") %>%
  left_join(
    as.data.frame(filter_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "FILTER"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(FILTER == "PASS", !is.na(DP))

# GB alleles (bp diff vs ref), PASS only
gb_long <- as.data.frame(gb_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "GB") %>%
  left_join(
    as.data.frame(filter_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "FILTER"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(FILTER == "PASS", !is.na(GB), GB != ".") %>%
  mutate(
    al1 = suppressWarnings(as.integer(str_extract(GB, "^-?[0-9]+"))),
    al2 = suppressWarnings(as.integer(str_extract(GB, "(?<=\\|)-?[0-9]+")))
  ) %>%
  pivot_longer(c(al1, al2), values_to = "bp_diff") %>%
  filter(!is.na(bp_diff)) %>%
  left_join(info_df %>% select(Loco, PERIOD), by = "Loco")
```

## D3-7. Global descriptive statistics

```{r d3_stats_globais}
tabela_global <- bind_rows(
  resumo_campo(dp_mat, "DP"),
  resumo_campo(q_mat, "Q"),
  resumo_campo(pq_mat, "PQ"),
  resumo_campo(gldiff_mat, "GLDIFF"),
  resumo_campo(dsnp_mat, "DSNP"),
  resumo_campo(dstut_mat, "DSTUTTER"),
  resumo_campo(dflank_mat, "DFLANKINDEL"),
  resumo_campo(ab_mat, "AB"),
  resumo_campo(fs_mat, "FS"),
  resumo_campo(dab_mat, "DAB"),
  resumo_campo(het_mat, "HET"),
  resumo_campo(hwep_mat, "HWEP"),
)

knitr::kable(tabela_global, digits = 4)

write.csv(tabela_global, "tabela_global_D3.csv", row.names = FALSE)
```

## D3-8. Statistics per locus

```{r d3_stats_por_loco}
stats_por_loco <- tibble(
  Loco             = loco_ids,
  CHROM            = vcf@fix[, "CHROM"],
  POS              = as.integer(vcf@fix[, "POS"]),
  DP_mean          = rowMeans(dp_mat,     na.rm = TRUE),
  DP_min           = apply(dp_mat, 1, min, na.rm = TRUE),
  DP_max           = apply(dp_mat, 1, max, na.rm = TRUE),
  Q_mean           = rowMeans(q_mat,      na.rm = TRUE),
  Q_min            = apply(q_mat, 1, min, na.rm = TRUE),
  PQ_mean          = rowMeans(pq_mat,     na.rm = TRUE),
  GLDIFF_mean      = rowMeans(gldiff_mat, na.rm = TRUE),
  GLDIFF_min       = apply(gldiff_mat, 1, min, na.rm = TRUE),
  DSTUTTER_mean    = rowMeans(dstut_mat,  na.rm = TRUE),
  DSTUTTER_max     = apply(dstut_mat, 1, max, na.rm = TRUE),
  DFLANKINDEL_mean = rowMeans(dflank_mat, na.rm = TRUE),
  AB_mean          = rowMeans(ab_mat,     na.rm = TRUE),
  AB_min           = apply(ab_mat, 1, min, na.rm = TRUE),
  FS_mean          = rowMeans(fs_mat,     na.rm = TRUE),
  FS_min           = apply(fs_mat, 1, min, na.rm = TRUE),
  N_PASS           = apply(filter_mat, 1,
                           function(x) sum(x == "PASS",  na.rm = TRUE)),
  N_NOCALL         = apply(filter_mat, 1,
                           function(x) sum(grepl("NOCALL|HipSTR", x),
                                           na.rm = TRUE)),
  N_NA             = rowSums(is.na(filter_mat)),
  Perc_PASS        = round(apply(filter_mat, 1,
                                 function(x) mean(x == "PASS",
                                                  na.rm = TRUE)) * 100, 1)
) %>%
  mutate(across(where(is.double), ~ round(.x, 4)))

cat("Distribution of % PASS calls per locus:\n")
summary(stats_por_loco$Perc_PASS)

knitr::kable(head(stats_por_loco, 10))
```

## D3-9. Statistics per sample

```{r d3_stats_por_amostra}
stats_por_amostra <- tibble(
  Amostra       = samples,
  DP_mean       = colMeans(dp_mat,     na.rm = TRUE),
  DP_min        = apply(dp_mat, 2, min, na.rm = TRUE),
  DP_max        = apply(dp_mat, 2, max, na.rm = TRUE),
  Q_mean        = colMeans(q_mat,      na.rm = TRUE),
  Q_min         = apply(q_mat, 2, min, na.rm = TRUE),
  GLDIFF_mean   = colMeans(gldiff_mat, na.rm = TRUE),
  DSTUTTER_mean = colMeans(dstut_mat,  na.rm = TRUE),
  AB_mean       = colMeans(ab_mat,     na.rm = TRUE),
  FS_mean       = colMeans(fs_mat,     na.rm = TRUE),
  N_PASS        = apply(filter_mat, 2,
                        function(x) sum(x == "PASS",  na.rm = TRUE)),
  N_NOCALL      = apply(filter_mat, 2,
                        function(x) sum(grepl("NOCALL|HipSTR", x),
                                        na.rm = TRUE)),
  Perc_PASS     = round(apply(filter_mat, 2,
                               function(x) mean(x == "PASS",
                                                na.rm = TRUE)) * 100, 1)
) %>%
  mutate(across(where(is.double), ~ round(.x, 4))) %>%
  arrange(desc(Perc_PASS))

knitr::kable(stats_por_amostra)
```

## D3-10. Filter distribution

```{r d3_filtros_freq}
todos_filtros <- as.vector(filter_mat)
todos_filtros[is.na(todos_filtros)] <- "NA"

contagem_filtros <- table(todos_filtros) %>%
  as.data.frame() %>%
  arrange(desc(Freq)) %>%
  rename(Categoria = todos_filtros) %>%
  mutate(Percentual = round(Freq / sum(Freq) * 100, 2))

knitr::kable(contagem_filtros)

write.csv(contagem_filtros, "contagem_filt_D3.csv")
```

## D3-11. GB variation (bp diff vs. reference)

```{r d3_gb_stats}
gb_vals_mask <- filter_mat == "PASS"
gb_vals_mask[is.na(gb_vals_mask)] <- FALSE

extrair_gb_num <- function(gb_str) {
  if (is.na(gb_str) || gb_str == ".") return(NA_real_)
  suppressWarnings(as.numeric(str_split(gb_str, "\\|")[[1]]))
}

gb_nums <- unlist(lapply(gb_mat[gb_vals_mask], extrair_gb_num))
gb_nums <- gb_nums[!is.na(gb_nums)]

tibble(
  Estatistica = c("Min", "Q25", "Median", "Mean", "Q75", "Max", "SD"),
  Valor_bp    = c(
    min(gb_nums),
    quantile(gb_nums, 0.25),
    median(gb_nums),
    round(mean(gb_nums), 2),
    quantile(gb_nums, 0.75),
    max(gb_nums),
    round(sd(gb_nums), 2)
  )
) %>% knitr::kable()
```

## D3-12. CSV table export

```{r d3_exportar_csv}
write_csv(tabela_global,     "d3_stats_global_campos_FORMAT.csv")
write_csv(stats_por_loco,    "d3_stats_por_loco.csv")
write_csv(stats_por_amostra, "d3_stats_por_amostra.csv")
write_csv(contagem_filtros,  "d3_distribuicao_filtros.csv")

cat("Files generated:\n",
    "  d3_stats_global_campos_FORMAT.csv\n",
    "  d3_stats_por_loco.csv\n",
    "  d3_stats_por_amostra.csv\n",
    "  d3_distribuicao_filtros.csv\n")
```

## D3-13. Figure 1 — Filters by sample

```{r d3_fig1, fig.width=10, fig.height=12}
filter_long <- as.data.frame(filter_mat) %>%
  pivot_longer(everything(), names_to = "Amostra", values_to = "FILTER") %>%
  filter(!is.na(FILTER)) %>%
  mutate(
    Categoria = case_when(
      FILTER == "PASS"   ~ "PASS",
      FILTER == "NOCALL" ~ "NOCALL",
      str_detect(FILTER, "MinDepth") ~ paste0("MinDepth (",
                                               str_extract(FILTER, "[0-9]+$"),
                                               ")"),
      TRUE ~ FILTER
    ),
    Categoria = fct_infreq(Categoria)
  ) %>%
  count(Amostra, Categoria) %>%
  group_by(Amostra) %>%
  mutate(pct = n / sum(n) * 100)

cores_filter <- c(
  "PASS"   = "#009E73",
  "NOCALL" = "#999999",
  setNames(
    colorRampPalette(c("#F0E442", "#D55E00"))(7),
    paste0("MinDepth (", 1:7, ")")
  )
)

p1a <- ggplot(filter_long,
              aes(x = reorder(Amostra, -pct * (Categoria == "PASS")),
                  y = pct, fill = Categoria)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = cores_filter, name = "Filter") +
  scale_y_continuous(expand = c(0, 0),
                     labels = label_percent(scale = 1)) +
  labs(
    title    = "A  Filter composition by sample",
    subtitle = "Proportion of each filter category across genotyped loci",
    x        = "Sample",
    y        = "% of loci"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7.5))

p1b <- perc_pass_amostra %>%
  arrange(perc_pass) %>%
  mutate(Amostra = factor(Amostra, levels = Amostra)) %>%
  ggplot(aes(x = perc_pass, y = Amostra)) +
  geom_segment(aes(xend = 0, yend = Amostra),
               color = "grey70", linewidth = 0.5) +
  geom_point(aes(color = perc_pass), size = 3.5) +
  scale_color_gradient2(
    low = "#D55E00", mid = "#E69F00", high = "#009E73",
    midpoint = 70, guide = "none"
  ) +
  geom_vline(xintercept = 70, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  scale_x_continuous(labels = label_percent(scale = 1),
                     limits = c(0, 100)) +
  labs(
    title    = "B  PASS call rate by sample",
    subtitle = "Dashed line = 70%",
    x        = "% loci with PASS",
    y        = NULL
  ) +
  theme(axis.text.y = element_text(size = 7.5))

fig1_d3 <- p1a / p1b + plot_layout(heights = c(1.4, 1))
fig1_d3
```

## D3-14. Figure 2 — Read depth (DP)

```{r d3_fig2, fig.width=10, fig.height=10}
p2a <- dp_pass %>%
  ggplot(aes(x = DP)) +
  geom_histogram(binwidth = 1, fill = "#0072B2", color = "white",
                 linewidth = 0.15, alpha = 0.9) +
  geom_vline(xintercept = median(dp_pass$DP), linetype = "dashed",
             color = "#D55E00", linewidth = 0.7) +
  annotate("text", x = median(dp_pass$DP) + 1.5, y = Inf,
           label = paste0("Median = ", median(dp_pass$DP)),
           hjust = 0, vjust = 1.5, size = 3, color = "#D55E00") +
  scale_x_continuous(breaks = c(8, 10, 15, 20, 30, 40, 50, 64)) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  labs(
    title    = "A  Distribution of read depth (DP)",
    subtitle = "Genotypes with PASS; n = all samples × loci",
    x        = "DP (reads per genotype)",
    y        = "Frequency"
  )

p2b <- dp_pass %>%
  mutate(Amostra = reorder(Amostra, DP, FUN = median)) %>%
  ggplot(aes(x = Amostra, y = DP)) +
  geom_boxplot(fill = "#56B4E9", color = "grey30", outlier.size = 0.6,
               outlier.alpha = 0.4, linewidth = 0.45, width = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 18,
               size = 2.2, color = "#D55E00") +
  scale_y_continuous(breaks = c(8, 10, 15, 20, 30, 40, 50, 64)) +
  labs(
    title    = "B  DP by sample (PASS genotypes)",
    subtitle = "Orange diamond = mean; boxes sorted by median",
    x        = "Sample",
    y        = "DP (reads)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7.5))

fig2_d3 <- p2a / p2b
fig2_d3
```

## D3-15. Figure 3 — Genotype quality (Q and GLDIFF)

```{r d3_fig3, fig.width=11, fig.height=10}
q_long <- as.data.frame(q_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "Q") %>%
  left_join(
    as.data.frame(filter_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "FILTER"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(FILTER == "PASS", !is.na(Q))

gldiff_long <- as.data.frame(gldiff_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "GLDIFF") %>%
  left_join(
    as.data.frame(filter_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "FILTER"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(FILTER == "PASS", !is.na(GLDIFF))

p3a <- q_long %>%
  ggplot(aes(x = Q)) +
  geom_histogram(binwidth = 0.005, boundary = 0.0025, fill = "#009E73", color = "white",
                 linewidth = 0.1, alpha = 0.9) +
  scale_x_continuous(limits = c(0.88, 1.005),
                     breaks = seq(0.88, 1.00, by = 0.02)) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  labs(
    title    = "A  Posterior probability of the genotype (Q)",
    subtitle = "PASS genotypes; values concentrated close to 1.0",
    x        = "Q (posterior probability)",
    y        = "Frequency"
  )

p3b <- gldiff_long %>%
  ggplot(aes(x = GLDIFF)) +
  geom_histogram(bins = 60, fill = "#CC79A7", color = "white",
                 linewidth = 0.1, alpha = 0.9) +
  geom_vline(xintercept = 2, linetype = "dashed",
             color = "grey40", linewidth = 0.6) +
  annotate("text", x = 3, y = Inf, label = "GLDIFF = 2",
           hjust = 0, vjust = 1.5, size = 3, color = "grey30") +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100),
                labels = c(1, 2, 5, 10, 20, 50, 100)) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  labs(
    title    = "B  Likelihood difference between genotypes (GLDIFF)",
    subtitle = "Log10 scale; higher values indicate greater confidence in the call",
    x        = "GLDIFF (log10)",
    y        = "Frequency"
  )

set.seed(42)
qgl_joined <- q_long %>%
  inner_join(gldiff_long, by = c("Loco", "Amostra")) %>%
  sample_n(min(5000, n()))

p3c <- qgl_joined %>%
  ggplot(aes(x = GLDIFF, y = Q)) +
  geom_hex(bins = 50) +
  scale_fill_gradientn(
    colors = c("grey90", "#56B4E9", "#0072B2", "#000040"),
    name   = "N genotypes",
    labels = label_comma()
  ) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100)) +
  scale_y_continuous(limits = c(0.88, 1.005),
                     breaks = c(0.9, 0.95, 0.99, 1.0)) +
  labs(
    title    = "C  Q × GLDIFF (hexbin)",
    subtitle = "Relationship between posterior probability and relative confidence",
    x        = "GLDIFF (log10)",
    y        = "Q"
  )

fig3_d3 <- (p3a | p3b) / p3c
fig3_d3
```

## D3-16. Figure 4 — Heterozygosity and HWE per locus

```{r d3_fig4, fig.width=11, fig.height=10}
p4a <- info_df %>%
  filter(!is.na(HET)) %>%
  ggplot(aes(x = HET)) +
  geom_histogram(binwidth = 0.02, fill = "#E69F00", color = "white",
                 linewidth = 0.15, alpha = 0.9) +
  geom_vline(xintercept = mean(info_df$HET, na.rm = TRUE),
             linetype = "dashed", color = "#0072B2", linewidth = 0.7) +
  annotate("text",
           x     = mean(info_df$HET, na.rm = TRUE) + 0.02,
           y     = Inf,
           label = paste0("Mean = ",
                          round(mean(info_df$HET, na.rm = TRUE), 3)),
           hjust = 0, vjust = 1.5, size = 3, color = "#0072B2") +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title    = "A  Expected heterozygosity (HET) per locus",
    subtitle = "Calculated by HipSTR; dashed line = global mean",
    x        = "He",
    y        = "Number of loci"
  )

p4b <- info_df %>%
  filter(!is.na(HET), !is.na(PERIOD), PERIOD >= 2) %>%
  ggplot(aes(x = factor(PERIOD), y = HET, fill = factor(PERIOD))) +
  geom_violin(alpha = 0.75, color = "grey40", linewidth = 0.35,
              scale = "width", trim = TRUE) +
  geom_boxplot(width = 0.12, outlier.size = 0.7, outlier.alpha = 0.4,
               fill = "white", color = "grey30", linewidth = 0.4) +
  scale_fill_manual(values = cores_okabe[1:8], guide = "none") +
  labs(
    title    = "B  HET by repeat motif size",
    subtitle = "Violin + boxplot; PERIOD = motif length in bp",
    x        = "Period (bp)",
    y        = "He"
  )

fig4_d3 <- p4a / p4b + plot_layout(heights = c(1, 1.6))
fig4_d3
```

## D3-17. Figure 5 — Allelic variation (GB)

```{r d3_fig5, fig.width=10, fig.height=11}
p5a <- gb_long %>%
  ggplot(aes(x = bp_diff)) +
  geom_histogram(binwidth = 1, fill = "#0072B2", color = "white",
                 linewidth = 0.1, alpha = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#D55E00",
             linewidth = 0.7) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  labs(
    title    = "A  Distribution of allelic variation (GB)",
    subtitle = "Difference in bp of each called allele vs. reference allele",
    x        = "GB (bp diff. vs. reference)",
    y        = "Allele frequency"
  )

p5b <- gb_long %>%
  filter(PERIOD >= 2, PERIOD <= 9) %>%
  mutate(PERIOD_lab = paste0("PERIOD = ", PERIOD, " bp")) %>%
  ggplot(aes(x = bp_diff,
             y = fct_rev(factor(PERIOD_lab)),
             fill = factor(PERIOD))) +
  geom_density_ridges(
    alpha          = 0.75,
    color          = "grey40",
    linewidth      = 0.3,
    bandwidth      = 1.5,
    scale          = 1.1,
    rel_min_height = 0.01
  ) +
  scale_fill_manual(values = cores_okabe[1:8], guide = "none") +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey30", linewidth = 0.5) +
  labs(
    title    = "B  Allelic variation by repeat motif period (ridgeline)",
    subtitle = "Distribution of bp_diff conditioned on repeat period",
    x        = "GB (bp diff. vs. reference)",
    y        = NULL
  )

fig5_d3 <- p5a / p5b + plot_layout(heights = c(1, 1.6))
fig5_d3
```

## D3-18. Figure 6 — Artifacts: stutter, allele bias, and strand bias

```{r d3_fig6, fig.width=10, fig.height=10}
art_df <- tibble(
  Loco          = loco_ids,
  PERIOD        = info_df$PERIOD,
  DSTUTTER_mean = rowMeans(dstut_mat,  na.rm = TRUE),
  DFLANK_mean   = rowMeans(dflank_mat, na.rm = TRUE)
)

p6a <- art_df %>%
  filter(!is.na(PERIOD), PERIOD >= 2) %>%
  group_by(PERIOD) %>%
  summarise(
    pct_stutter = mean(DSTUTTER_mean > 0, na.rm = TRUE) * 100,
    n_locos     = n()
  ) %>%
  ggplot(aes(x = factor(PERIOD), y = pct_stutter, fill = factor(PERIOD))) +
  geom_col(color = "white", linewidth = 0.2, width = 0.7) +
  scale_fill_manual(values = cores_okabe[1:8], guide = "none") +
  scale_y_continuous(labels = label_percent(scale = 1), expand = c(0, 0)) +
  labs(
    title    = "A  % of loci with detected stutter (DSTUTTER > 0) by period",
    subtitle = "Stutter is an expected artifact in STRs, especially for short motifs",
    x        = "Period (bp)",
    y        = "% of loci"
  )

ab_long <- as.data.frame(ab_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "AB") %>%
  left_join(
    as.data.frame(filter_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "FILTER"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(FILTER == "PASS", !is.na(AB), AB < 0)

fs_long <- as.data.frame(fs_mat) %>%
  mutate(Loco = loco_ids) %>%
  pivot_longer(-Loco, names_to = "Amostra", values_to = "FS") %>%
  left_join(
    as.data.frame(filter_mat) %>%
      mutate(Loco = loco_ids) %>%
      pivot_longer(-Loco, names_to = "Amostra", values_to = "FILTER"),
    by = c("Loco", "Amostra")
  ) %>%
  filter(FILTER == "PASS", !is.na(FS), FS < 0)

bias_long <- bind_rows(
  ab_long %>% transmute(valor = AB, Tipo = "AB (allele bias)"),
  fs_long %>% transmute(valor = FS, Tipo = "FS (strand bias)")
)

p6b <- bias_long %>%
  ggplot(aes(x = valor, fill = Tipo)) +
  geom_histogram(binwidth = 0.05, alpha = 0.8, color = "white",
                 linewidth = 0.1, position = "identity") +
  geom_vline(
  data = data.frame(x_int = -2, tipo_linha = "dashed"),
  aes(xintercept = x_int, linetype = I(tipo_linha)),
  color = "grey30", linewidth = 0.5
) +
  scale_fill_manual(values = c("#E69F00", "#CC79A7"), name = NULL) +
  scale_y_continuous(labels = label_comma(), expand = c(0, 0)) +
  facet_wrap(~Tipo, scales = "free_y") +
  labs(
    title    = "B  Distribution of allele bias (AB) and strand bias (FS)",
    subtitle = "PASS heterozygous genotypes only; values = log10(p-value)",
    x        = "log10(bias p-value)",
    y        = "Frequency"
  ) +
  theme(legend.position = "none")

fig6_d3 <- p6a / p6b
fig6_d3
```

## D3-19. Figure 7 — DP heatmap by sample × chromosome

```{r d3_fig7, fig.width=9, fig.height=8}
dp_chr <- as.data.frame(dp_mat) %>%
  mutate(Loco = loco_ids, CHROM = vcf@fix[, "CHROM"]) %>%
  filter(str_detect(CHROM, "^Dal_")) %>%
  pivot_longer(-c(Loco, CHROM), names_to = "Amostra", values_to = "DP") %>%
  left_join(
    as.data.frame(filter_mat) %>%
      mutate(Loco = loco_ids, CHROM = vcf@fix[, "CHROM"]) %>%
      pivot_longer(-c(Loco, CHROM), names_to = "Amostra",
                   values_to = "FILTER"),
    by = c("Loco", "Amostra", "CHROM")
  ) %>%
  filter(FILTER == "PASS") %>%
  group_by(CHROM, Amostra) %>%
  summarise(DP_mean = mean(DP, na.rm = TRUE), .groups = "drop")

fig7_d3 <- dp_chr %>%
  mutate(
    CHROM   = factor(CHROM, levels = paste0("Dal_0", 1:8)),
    Amostra = reorder(Amostra, DP_mean)
  ) %>%
  ggplot(aes(x = CHROM, y = Amostra, fill = DP_mean)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradientn(
    colors = c("#f7fbff", "#c6dbef", "#6baed6", "#2171b5", "#08306b"),
    name   = "Mean DP",
    limits = c(8, NA)
  ) +
  geom_text(aes(label = round(DP_mean, 1)), size = 2.5, color = "white") +
  labs(
    title    = "Mean read depth (DP) by sample × chromosome",
    subtitle = "PASS loci only on main chromosomes (Dal_01–Dal_08)",
    x        = "Chromosome",
    y        = "Sample"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

fig7_d3
```

## D3-20. Physical map and locus density in the genome

```{r d3_fig_genomic_maps_motif, fig.width=16, fig.height=10}
# Load chromosome sizes
chrom_sizes <- read_tsv("chromosome_sizes.txt", 
                        col_names = c("chr", "size_bp"), 
                        show_col_types = FALSE)

total_chrom_size_bp <- sum(chrom_sizes$size_bp)

# Prepare data with PERIOD
map_data <- tibble(
  locus = loco_ids,
  chr = vcf@fix[, "CHROM"],
  pos = as.numeric(vcf@fix[, "POS"]),
  position_mb = pos / 1e6,
  period = info_df$PERIOD
) %>%
  filter(!is.na(pos), !is.na(period)) %>%
  mutate(
    motif_class = case_when(
      period == 1 ~ "Mono",
      period == 2 ~ "Di",
      period == 3 ~ "Tri",
      period == 4 ~ "Tetra",
      period == 5 ~ "Penta",
      period == 6 ~ "Hexa",
      period == 7 ~ "Hepta",
      period == 8 ~ "Octa",
      period == 9 ~ "Nona",
      TRUE ~ "Other"
    ),
    motif_class = factor(motif_class, 
                         levels = c("Mono", "Di", "Tri", "Tetra", "Penta", 
                                   "Hexa", "Hepta", "Octa", "Nona", "Other"))
  )

chrom_plot_data <- chrom_sizes %>%
  mutate(size_mb = size_bp / 1e6, chr = as.character(chr)) %>%
  filter(chr %in% unique(map_data$chr))

# Okabe-Ito palette for motifs
motif_colors <- c(
  "Mono"  = "#56B4E9", "Di"    = "#0072B2", "Tri"   = "#009E73",
  "Tetra" = "#F0E442", "Penta" = "#E69F00", "Hexa"  = "#D55E00",
  "Hepta" = "#CC79A7", "Octa"  = "#999999", "Nona"  = "#2C2C2A", 
  "Other" = "#CCCCCC"
)

# Legend with counts
motif_counts <- map_data %>%
  count(motif_class, name = "n") %>%
  mutate(
    pct = round((n / sum(n)) * 100, 1),
    label = paste0(motif_class, " (n=", format(n, big.mark = ","), 
                   ", ", pct, "%)")
  )

legend_labels <- setNames(motif_counts$label, motif_counts$motif_class)

cat("✓ Locus map prepared (D3 - PASS loci):\n",
    "  -", nrow(map_data), "loci mapped with PERIOD\n",
    "  -", length(unique(map_data$chr)), "chromosomes\n",
    "  - Distribution by motif:\n")
print(motif_counts)

# ---- PLOT 1: Physical Map by Motif ----
physical_map <- ggplot() +
  geom_rect(
    data = chrom_plot_data,
    aes(xmin = as.numeric(factor(chr)) - 0.38,
        xmax = as.numeric(factor(chr)) + 0.38,
        ymin = 0, ymax = size_mb),
    fill = "grey96", color = "grey70", linewidth = 0.4
  ) +
  geom_linerange(
    data = map_data,
    aes(x = chr, ymin = position_mb, ymax = position_mb + 0.08, 
        color = motif_class),
    linewidth = 16.0, alpha = 1.0
  ) +
  scale_color_manual(
    values = motif_colors, 
    labels = legend_labels,
    name = "Motif type",
    guide = guide_legend(ncol = 1)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.03)),
    breaks = seq(0, 700, by = 50),
    labels = function(x) paste0(x, " Mb")
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1, color = "grey20"),
    axis.text.y = element_text(size = 9, color = "grey20"),
    axis.title = element_text(size = 11, face = "bold", color = "grey10"),
    axis.line = element_line(color = "grey30", linewidth = 0.4),
    axis.ticks = element_line(color = "grey30", linewidth = 0.4),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.35),
    panel.border = element_rect(color = "grey75", linewidth = 0.5),
    legend.title = element_text(size = 9.5, face = "bold", color = "grey10"),
    legend.text = element_text(size = 8, color = "grey25"),
    legend.key.size = unit(0.35, "cm"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5,
                              color = "grey10", margin = margin(b = 6)),
    plot.subtitle = element_text(size = 9.5, color = "grey40", hjust = 0.5,
                                 margin = margin(b = 8)),
    plot.margin = margin(12, 12, 10, 12)
  ) +
  labs(
    title = "A  Physical map: STR loci by motif repeat period",
    subtitle = paste0(format(nrow(map_data), big.mark = ","), " loci across ",
                      length(unique(map_data$chr)), " chromosomes (PASS genotypes)"),
    x = "Chromosome",
    y = "Position (Mb)"
  )

# ---- PLOT 2: Density Heatmap (WITHOUT separation by motif) ----
window_mb <- 2.0

density_data <- map_data %>%
  mutate(bin = floor(position_mb / window_mb) * window_mb) %>%
  count(chr, bin, name = "n_loci") %>%
  left_join(chrom_plot_data %>% select(chr, size_mb), by = "chr") %>%
  group_by(chr) %>%
  do({
    bins_seq <- seq(0, floor(max(.$size_mb, na.rm = TRUE) / window_mb) * window_mb, 
                    by = window_mb)
    complete(., bin = bins_seq, fill = list(n_loci = 0))
  }) %>%
  filter(bin < size_mb) %>%
  mutate(
    tile_top = pmin(bin + window_mb, size_mb),
    tile_center = (bin + tile_top) / 2,
    tile_height = tile_top - bin
  ) %>%
  ungroup()

density_map <- ggplot() +
  geom_rect(
    data = chrom_plot_data,
    aes(xmin = as.numeric(factor(chr)) - 0.42,
        xmax = as.numeric(factor(chr)) + 0.42,
        ymin = 0, ymax = size_mb + 0.5),
    fill = "grey97", color = "grey65", linewidth = 0.4
  ) +
  geom_tile(
    data = density_data,
    aes(x = chr, y = tile_center, fill = n_loci, height = tile_height),
    width = 0.79, color = NA
  ) +
  scale_fill_gradientn(
    colors = c("white", "#B3D9FF", "#4D94FF", "#0052CC", "#001A4D"),
    name = "# loci",
    guide = guide_colorbar(barwidth = 0.7, barheight = 4, ticks = FALSE)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.03)),
    breaks = seq(0, 700, by = 50),
    labels = function(x) paste0(x, " Mb")
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1, color = "grey20"),
    axis.text.y = element_text(size = 9, color = "grey20"),
    axis.title = element_text(size = 11, face = "bold", color = "grey10"),
    axis.line = element_line(color = "grey30", linewidth = 0.4),
    axis.ticks = element_line(color = "grey30", linewidth = 0.4),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.35),
    panel.border = element_rect(color = "grey75", linewidth = 0.5),
    legend.title = element_text(size = 9.5, face = "bold", color = "grey10"),
    legend.text = element_text(size = 8, color = "grey25"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5,
                              color = "grey10", margin = margin(b = 6)),
    plot.subtitle = element_text(size = 9.5, color = "grey40", hjust = 0.5,
                                 margin = margin(b = 8)),
    plot.margin = margin(12, 12, 10, 12)
  ) +
  labs(
    title = "B  STR density heatmap across chromosomes",
    subtitle = paste0("Window size: ", window_mb, " Mb (PASS loci only)"),
    x = "Chromosome",
    y = "Position (Mb)"
  )

fig_genomic_d3 <- physical_map | density_map
fig_genomic_d3

# Export
ggsave("d3_physical_map_by_motif.png", plot = physical_map,
       width = 220, height = 160, units = "mm", dpi = 600, bg = "white")
ggsave("d3_density_heatmap.png", plot = density_map,
       width = 220, height = 160, units = "mm", dpi = 600, bg = "white")

cat("\n✓ Figures exported:\n",
    "  d3_physical_map_by_motif.png\n",
    "  d3_density_heatmap.png\n")
```

